\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{setspace}
\setstretch{1.1}
\newcolumntype{Y}{>{\raggedright\arraybackslash}X}
\newcommand{\taxon}[1]{\emph{#1}}
\newcommand{\dcstsep}{/\allowbreak}
\newcommand{\agpantibioticstatus}{\texttt{Antibiotic\_\allowbreak Status}}
\title{Supplementary Materials\\Inflammatory Bowel Disease-Associated Dominance States in the Human Gut Microbiome}
\author{Pawel Gajer and Jacques Ravel}
\date{}

\begin{document}
\maketitle

\section*{Supplementary Methods}

Outcome flags in the AGP discovery cohort were derived directly from the
\texttt{Phenotype} metadata field by fixed substring matching. The five
outcome variables analyzed in the main manuscript were IBS (\texttt{irritable bowel
syndrome}), IBD (\texttt{inflammatory bowel disease}), autoimmune disease
(\texttt{autoimmune}), acid reflux (\texttt{acid reflux}), and seasonal
allergies (\texttt{seasonal allergies}). For each outcome, samples
without the corresponding outcome-defining phrase served as controls.
These outcome flags were therefore not mutually exclusive across the
discovery cohort.

dCST construction was applied after the discovery feature-retention
filter described in the main manuscript. The labels should therefore be
read as retained-feature dominance states, not as claims about every raw
feature in the original table. Feature names use the deepest informative
SILVA annotation available for each retained feature, which is why the
tables include a mixture of species, genus, family, order, and unresolved
labels.

The AGP discovery phenotype labels are self-reported and do not provide a
uniform treatment-naive clinical IBD classification. IBD subtype, disease
activity, and detailed medication exposure were not consistently
available for adjusted discovery modeling. The coarser AGP
\agpantibioticstatus{} field was available for most age/sex/BMI
complete-case samples and was used in sensitivity analyses that either
added antibiotic status as a categorical covariate or excluded samples
reporting antibiotic use in the previous week, month, or six months. The
AGP associations are therefore interpreted as phenotype-screening
results and are paired with external validation rather than treated as
clinical mechanistic estimates on their own.

The adjusted frequentist label-level analyses used logistic regression
with age, sex, and BMI as covariates. The adjusted Bayesian follow-up
used \texttt{arm::bayesglm} with the same covariate structure and the
package's default weakly informative prior. For each label we summarize
the adjusted Bayesian fit by the exponentiated coefficient estimate, an
approximate 95\% credible interval based on the posterior standard
error, and the posterior probability that the label coefficient exceeds
0. Unadjusted Bayesian prevalence contrasts were estimated separately
with a beta-binomial model using a Jeffreys prior,
\(\mathrm{Beta}(0.5,0.5)\). BH-adjusted frequentist results are treated
as the primary screen-level evidence, whereas Bayesian estimates are used
as complementary stabilized summaries for sparse labels.

External validation was evaluated in two complementary modes. In the
rebuilt-cohort mode, each external count matrix was filtered and
analyzed independently with the same absorb workflow used for the AGP
discovery cohort. In the direct AGP-derived label-transfer mode,
external samples were mapped onto the AGP-derived absorb dCST hierarchy after aligning the
external taxa to the AGP retained-feature dictionary. The first mode
tests whether disease-associated dominance structure is reproducible
when rebuilt within a cohort; the second tests portability of the
AGP-defined label dictionary itself. Because some external datasets have
repeated samples per subject, we also repeated leading validation tests
after retaining one detected sample per subject. For direct AGP-derived
label transfer, we additionally summarized case/control mapping rates by
depth so that label-transfer associations could be interpreted alongside
mapping balance.

\section*{Supplementary Results}

\noindent\textbf{Acid reflux.} Acid reflux was the clearest additional
non-IBD gastrointestinal branch in the AGP screen. At depth 1 it was
associated with enrichment of \taxon{Bacteroides} (adjusted OR 1.28, q
= 1.75e-05) together with depletion of \taxon{Streptococcus} and
\taxon{Staphylococcus}. The signal then sharpened into deeper
\taxon{Bacteroides}-centered dominance-lineages enriched for
\taxon{Parabacteroides} and \taxon{Alistipes}, including
\taxon{Bacteroides}\dcstsep\taxon{Parabacteroides}\dcstsep Lachnospiraceae
(adjusted OR 2.64, q = 0.00419). Bayesian follow-up retained support
for numerous depth-4 descendants after the frequentist signal thinned.
Because reflux symptoms are self-reported and likely influenced by diet,
medication, and upper-gastrointestinal factors, we regard this branch as
hypothesis-generating rather than disease-specific.

\medskip

\noindent\textbf{Seasonal allergies.} Seasonal allergies provided a
complementary immune-linked example outside the core inflammatory
phenotypes. The dominant pattern was depletion of
\taxon{Prevotella\_9}-linked states together with enrichment of
enterobacterial dominance-lineages. At depth 1,
\taxon{Prevotella\_9} was depleted (adjusted OR 0.67, q = 2.77e-08),
whereas \taxon{Escherichia-Shigella} was modestly enriched. At depth 2,
the strongest enriched state was
\taxon{Escherichia-Shigella}\dcstsep\taxon{Bacteroides} (adjusted OR
1.24, q = 7.42e-06), while
\taxon{Prevotella\_9}\dcstsep\taxon{Bacteroides} was depleted (adjusted
OR 0.64, q = 0.000215). Bayesian follow-up preserved this paired
pattern at greater depth, but we interpret the branch as exploratory
rather than as a fully validated disease axis.

\section*{Supplementary Tables}

\noindent\textbf{Supplementary Table S1.} Definitions and analysis sample
sizes for the five AGP phenotype indicators analyzed in the main
manuscript.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{} l Y r r r r @{}}
\toprule
Phenotype & AGP metadata phrase used for case coding & Cases & Controls & Complete-case cases & Complete-case controls \\
\midrule
IBS & \texttt{irritable bowel syndrome} & 3,105 & 27,185 & 2,551 & 20,647 \\
IBD & \texttt{inflammatory bowel disease} & 1,166 & 29,124 & 993 & 22,205 \\
Autoimmune disease & \texttt{autoimmune} & 2,872 & 27,418 & 2,426 & 20,772 \\
Acid reflux & \texttt{acid reflux} & 2,859 & 27,431 & 2,279 & 20,919 \\
Seasonal allergies & \texttt{seasonal allergies} & 9,493 & 20,797 & 8,284 & 14,914 \\
\bottomrule
\end{tabularx}
\end{center}

\newpage

\noindent\textbf{Supplementary Table S2.} Growth of the absorb dCST hierarchy and
divergence from the pure hierarchy.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} r r r r Y r r @{}}
\toprule
Depth & \(n_0\) & Named states & Median size & Largest state (size) & Assignments changed & Changed (\%) \\
\midrule
1 & 50 & 40 & 182 & \taxon{Bacteroides} (10,599) & 1,162 & 3.8 \\
2 & 25 & 153 & 88 & \shortstack[l]{\taxon{Escherichia-Shigella}\dcstsep\\\taxon{Bacteroides} (3,728)} & 5,593 & 18.5 \\
3 & 12 & 479 & 39 & \shortstack[l]{\taxon{Bacteroides}\dcstsep\taxon{Faecalibacterium}\dcstsep\\Lachnospiraceae (1,278)} & 15,747 & 52.0 \\
4 & 10 & 744 & 33 & \shortstack[l]{Eukaryota\dcstsep Unassigned\dcstsep\\Bacteria\dcstsep Mitochondria (332)} & 21,699 & 71.6 \\
\bottomrule
\end{tabularx}
\end{center}

\noindent Some high-depth absorb labels include low-resolution or
potential non-gut annotations such as \emph{Eukaryota},
\emph{Unassigned}, Bacteria, and Mitochondria. These labels are retained
for transparent accounting of the unsupervised hierarchy, but they are
not used as biological anchors for the IBD interpretation in the main
text.

\bigskip

\noindent\textbf{Supplementary Table S3.} Selected IBD absorb
dominance-lineages
highlighting agreement between frequentist and Bayesian follow-up.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{} Y r r Y Y @{}}
\toprule
Label & Depth & \(n\) & Frequentist adjusted OR (95\% CI) & Bayesian adjusted OR (95\% CrI) \\
\midrule
\taxon{Bacteroides} & 1 & 10,599 & 1.54 (1.36--1.76) & 1.54 (1.36--1.76) \\
Enterobacterales & 1 & 103 & 3.23 (1.73--6.04) & 3.13 (1.68--5.84) \\
\taxon{Morganella} & 1 & 56 & not included in primary adjusted screen & 18.99 (10.15--35.53) \\
\taxon{Proteus} & 1 & 51 & not included in primary adjusted screen & 3.40 (1.40--8.27) \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae & 2 & 1,643 & 3.21 (2.67--3.86) & 3.20 (2.66--3.85) \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} & 3 & 176 & 8.56 (5.92--12.36) & 8.44 (5.85--12.19) \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes}\dcstsep\taxon{Faecalibacterium} & 4 & 89 & 5.53 (3.10--9.87) & 5.36 (3.01--9.55) \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S4A.} Rebuilt-cohort external
validation summary. Each external cohort was analyzed independently with
the same absorb-policy dCST workflow used for the discovery cohort.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{} l l Y Y @{}}
\toprule
Cohort & Comparison & Best rebuilt signal & Interpretation \\
\midrule
Halfvarson 2017 & IBD vs healthy & \taxon{Segatella} sp.\ (depth 1, q = 3.0e-09) & Strong rebuilt inflammatory replication. \\
Halfvarson 2017 & Crohn vs healthy & \taxon{Segatella} sp.\ (depth 1, q = 2.6e-05) & Strong rebuilt Crohn replication. \\
Halfvarson 2017 & UC vs healthy & \taxon{Bacteroides} \emph{dorei} (depth 1, q = 3.26e-09) & Strong rebuilt UC replication. \\
HMP2 / IBDMDB & IBD vs healthy & \taxon{Bacteroides}\dcstsep\taxon{Faecalibacterium} (depth 2, q = 0.24) & Directional only in rebuilt mode. \\
Gevers 2014 & Crohn vs healthy & \emph{Faecalibacterium prausnitzii} (depth 1, q = 0.305) & No corrected rebuilt signal. \\
PRJEB84421 & OFG vs healthy & \taxon{Faecalibacterium}\dcstsep\taxon{Bacteroides} (depth 2, q = 0.019) & Supportive rebuilt inflammatory signal. \\
Jacobs 2023 IBS 250bp & IBS vs healthy & \emph{Faecalibacterium prausnitzii}\dcstsep\emph{Bacteroides dorei}\dcstsep\emph{Bacteroides ovatus} (depth 4, q = 0.564) & Directional at most. \\
\bottomrule
\end{tabularx}
\end{center}

\medskip

\noindent\textbf{Supplementary Table S4B.} Direct AGP-derived label-transfer
validation summary. External samples were mapped onto the AGP-derived absorb dCST
hierarchy, testing portability of the AGP label dictionary.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{} l l Y r Y @{}}
\toprule
Cohort & Comparison & Best AGP-derived label-transfer signal & Depth-2 coverage (\%) & Interpretation \\
\midrule
Halfvarson 2017 & IBD vs healthy & \taxon{Bacteroides} (depth 1, q = 6.8e-05) & 92.3 & Strong transferred pooled-IBD signal. \\
Halfvarson 2017 & Crohn vs healthy & \taxon{Faecalibacterium} (depth 1, q = 6.5e-04) & 87.8 & Strong transferred Crohn signal. \\
Halfvarson 2017 & UC vs healthy & \taxon{Bacteroides} (depth 1, q = 2.96e-07) & 94.0 & Strong transferred UC signal. \\
HMP2 / IBDMDB & IBD vs healthy & \taxon{Bacteroides}\dcstsep\taxon{Faecalibacterium} (depth 2, q = 0.027) & 98.8 & Weaker but positive pooled-IBD transfer after explicit case coding. \\
Gevers 2014 & Crohn vs healthy & \taxon{Bacteroides}\dcstsep\taxon{Dialister} (depth 2, q = 0.972) & 77.8 & Samples map into the hierarchy, but no corrected transferred signal. \\
PRJEB84421 & OFG vs healthy & \taxon{Faecalibacterium} (depth 1, q = 0.282) & 81.6 & No matched AGP-derived label-transfer signal. \\
Jacobs 2023 IBS 250bp & IBS vs healthy & \taxon{Bacteroides}\dcstsep\taxon{Alistipes} (depth 2, q = 0.722) & 93.7 & IBS validation remains inconclusive. \\
\bottomrule
\end{tabularx}
\end{center}

\medskip

\noindent\textbf{Supplementary Table S4C.} External-validation cohort
context for the main comparisons summarized in Table 1.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} l Y Y @{}}
\toprule
Cohort & Samples and comparison labels & Context and caveat \\
\midrule
Halfvarson 2017 & 533 pooled IBD-vs-healthy samples (479/54); subset contrasts used 200/54 Crohn-vs-healthy and 279/54 UC-vs-healthy samples. & Stool, longitudinal IBD cohort; pooled comparison contained 118 detected subjects, 99 with repeated samples. \\
HMP2 / IBDMDB & 166 pooled IBD-vs-healthy samples after explicit case recoding (121/45). & Mucosal gut-site IBDMDB samples; 81 detected subjects, 67 with repeated samples. \\
PRJEB84421 & 49 OFG-vs-healthy samples (29/20). & Stool-based pediatric-onset orofacial granulomatosis comparison; no repeated subjects detected. \\
Gevers 2014 & 396 Crohn-vs-healthy samples (363/33). & Pediatric stool Crohn disease comparison; 363 detected subjects, 8 with repeated samples. \\
Jacobs 2023 IBS 250bp & 239 IBS-vs-healthy baseline stool samples (85/154). & 250bp IBS validation subset; no repeated subjects detected. \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S5.} Top depth-1 absorb-policy dCSTs in the full
AGP cohort.

\begin{center}
\footnotesize
\begin{tabular}{@{}lllll@{}}
\toprule
Rank & dCST & n & Share (\%) & Cumulative (\%) \\
\midrule
1 & \taxon{Bacteroides} & 10599 & 35.0 & 35.0 \\
2 & \taxon{Escherichia-Shigella} & 6913 & 22.8 & 57.8 \\
3 & \taxon{Prevotella\_9} & 1926 & 6.4 & 64.2 \\
4 & Enterobacteriaceae & 1596 & 5.3 & 69.4 \\
5 & \taxon{Faecalibacterium} & 1405 & 4.6 & 74.1 \\
6 & \taxon{Streptococcus} & 881 & 2.9 & 77.0 \\
7 & \taxon{Pseudomonas} & 690 & 2.3 & 79.3 \\
8 & \taxon{Staphylococcus} & 641 & 2.1 & 81.4 \\
9 & \taxon{Prevotella\_7} & 498 & 1.6 & 83.0 \\
10 & Lachnospiraceae & 480 & 1.6 & 84.6 \\
11 & Unassigned & 464 & 1.5 & 86.1 \\
12 & \taxon{Akkermansia} & 348 & 1.1 & 87.3 \\
\bottomrule
\end{tabular}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S6.} Selected absorb-policy dCST
dominance-lineage follow-up summary across the main-text and
supplementary phenotype branches.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} l Y Y Y @{}}
\toprule
Phenotype & Strongest depth-1 adjusted signal & Strongest deeper adjusted signal & Bayesian value added \\
\midrule
IBD & \taxon{Bacteroides} enriched (adjusted OR 1.54, q = 8.1e-10) & \taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} enriched at depth 3 (adjusted OR 8.56, q = 1.5e-28) & Bayesian models highlighted rare enterobacterial states, especially \taxon{Morganella}, that were outside the primary adjusted screen. \\
IBS & \taxon{Escherichia-Shigella} enriched (adjusted OR 1.24, q = 0.002) & \taxon{Escherichia-Shigella}\dcstsep\taxon{Bacteroides}\dcstsep Lachnospiraceae enriched at depth 3 (adjusted OR 1.78, q = 0.018) & Bayesian models retained support for several depth-4 dominance-lineages even when q-values no longer crossed 0.05. \\
Autoimmune & \taxon{Bacteroides} enriched (adjusted OR 1.40, q = 1.6e-12) & \taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} enriched at depth 3 (adjusted OR 5.52, q = 3.9e-21) & Bayesian models supported additional depth-4 \taxon{Bacteroides}-centered dominance-lineages beyond the frequentist q-significant set. \\
Acid reflux & \taxon{Bacteroides} enriched (adjusted OR 1.28, q = 1.75e-05) & \taxon{Bacteroides}\dcstsep\taxon{Parabacteroides}\dcstsep Lachnospiraceae enriched at depth 3 (adjusted OR 2.64, q = 0.00419) & Bayesian models retained support for numerous depth-4 \taxon{Bacteroides}-centered reflux dominance-lineages after the frequentist signal thinned. \\
Seasonal allergies & \taxon{Prevotella\_9} depleted (adjusted OR 0.67, q = 2.77e-08) & \taxon{Escherichia-Shigella}\dcstsep\taxon{Bacteroides} enriched at depth 2 (adjusted OR 1.24, q = 7.42e-06) & Bayesian models clarified the paired pattern of depleted \taxon{Prevotella}-linked states and enriched enterobacterial dominance-lineages at depths 3 and 4. \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S7.} IBD sensitivity analysis after
excluding non-IBD controls carrying any other AGP phenotype string
(IBS, autoimmune disease, acid reflux, or seasonal allergies). The
analysis retained all 1,166 IBD cases and 16,175 cleaner controls.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{} Y r Y Y Y @{}}
\toprule
Label & Depth & Primary analysis & Cleaner-control sensitivity & Interpretation \\
\midrule
\taxon{Bacteroides} & 1 & 511/10,088 cases/controls; adjusted OR 1.54, q = 8.1e-10. & 511/5,409; adjusted OR 1.59, q = 1.1e-10. & Stable broad IBD signal. \\
\taxon{Morganella} & 1 & 22/34; not included in the primary adjusted frequentist screen; Bayesian adjusted OR 18.99. & Below support threshold after control exclusion. & Rare-state signal remains exploratory. \\
\taxon{Proteus} & 1 & 6/45; not included in the primary adjusted frequentist screen; Bayesian adjusted OR 3.40. & Below support threshold after control exclusion. & Rare-state signal remains exploratory. \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae & 2 & 163/1,480; adjusted OR 3.21, q = 2.6e-33. & 163/791; adjusted OR 3.15, q = 1.5e-28. & Stable deeper IBD dCST dominance-lineage. \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} & 3 & 49/127; adjusted OR 8.56, q = 1.5e-28. & 49/55; adjusted OR 10.15, q = 6.2e-22. & Stable deeper IBD dCST dominance-lineage. \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes}\dcstsep\taxon{Faecalibacterium} & 4 & 16/73; adjusted OR 5.53, q = 1.1e-07. & 16/35; frequentist model skipped; Bayesian adjusted OR 5.68. & Direction retained, but depth-4 support becomes sparse. \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S8.} Retained-feature profiles for
selected IBD dCST dominance-lineages. Values are median retained-feature
relative abundances (\%) within samples assigned to each label; the IBD-case
median is shown in parentheses. This table clarifies that a dCST dominance-lineage
is an ordered retained-feature context rather than a claim that each
component is globally enriched in IBD.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} Y r r r r r r @{}}
\toprule
Label & Depth & IBD/control & \taxon{Bacteroides} & Lachnospiraceae & \taxon{Alistipes} & \taxon{Faecalibacterium} \\
\midrule
\taxon{Bacteroides} & 1 & 511/10,088 & 32.0 (24.3) & 4.7 (7.1) & 2.5 (4.3) & 5.7 (4.9) \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae & 2 & 163/1,480 & 31.8 (20.4) & 10.1 (10.6) & 2.2 (5.3) & 5.3 (4.9) \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} & 3 & 49/127 & 30.9 (18.2) & 9.6 (10.5) & 6.6 (7.9) & 3.7 (4.9) \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes}\dcstsep\taxon{Faecalibacterium} & 4 & 16/73 & 31.2 (21.9) & 9.4 (10.1) & 6.7 (8.4) & 4.5 (6.2) \\
\bottomrule
\end{tabularx}
\end{center}

\newpage

\noindent\textbf{Supplementary Table S9.} Antibiotic-status sensitivity
for selected IBD dCST dominance-lineages. Values are adjusted odds
ratios and selected-label BH-adjusted q-values. The three columns compare
the age/sex/BMI complete-case model, the same model with coarse AGP
antibiotic status added as a categorical covariate, and a
model excluding samples reporting antibiotic use in the previous week,
month, or six months.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} Y r Y Y Y @{}}
\toprule
Label & Depth & Age/sex/BMI & Plus antibiotic status & Excluding recent antibiotics \\
\midrule
\taxon{Bacteroides} & 1 & OR 1.54; q = 6.7e-11 & OR 1.58; q = 9.5e-12 & OR 1.69; q = 1.5e-12 \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae & 2 & OR 3.21; q = 1.4e-34 & OR 3.26; q = 3.4e-35 & OR 3.44; q = 3.5e-33 \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} & 3 & OR 8.56; q = 8.0e-30 & OR 8.64; q = 1.4e-29 & OR 10.04; q = 2.6e-30 \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes}\dcstsep\taxon{Faecalibacterium} & 4 & OR 5.53; q = 8.5e-09 & OR 5.67; q = 7.0e-09 & OR 6.51; q = 2.3e-09 \\
\taxon{Morganella} & 1 & OR 19.77; q = 3.2e-20 & OR 19.20; q = 4.1e-19 & OR 19.42; q = 2.0e-18 \\
\taxon{Proteus} & 1 & OR 3.63; q = 0.0047 & OR 3.75; q = 0.0040 & OR 6.46; q = 0.00048 \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S10.} First-sample-per-subject
sensitivity for external validation. Values show the sample-level
BH-adjusted q-value followed by the corresponding q-value after retaining
one detected sample per subject for the same leading label.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} l l Y Y Y @{}}
\toprule
Cohort & Comparison & Rebuilt-cohort q & AGP-derived label-transfer q & Interpretation \\
\midrule
Halfvarson 2017 & IBD vs healthy & 3.0e-09 \(\rightarrow\) 0.219 & 6.8e-05 \(\rightarrow\) 0.097 & Strong sample-level support; subject-level pooled transfer becomes suggestive. \\
Halfvarson 2017 & Crohn vs healthy & 2.6e-05 \(\rightarrow\) 0.102 & 6.5e-04 \(\rightarrow\) 0.466 & Sample-level support attenuates after repeated-sample reduction. \\
Halfvarson 2017 & UC vs healthy & 3.3e-09 \(\rightarrow\) 0.075 & 3.0e-07 \(\rightarrow\) 0.0029 & Transferred UC label remains significant at subject level. \\
HMP2 / IBDMDB & IBD vs healthy & 0.240 \(\rightarrow\) 0.075 & 0.027 \(\rightarrow\) 0.012 & Selective transferred depth-2 signal persists at subject level. \\
Gevers 2014 & Crohn vs healthy & 0.305 \(\rightarrow\) 0.291 & 0.972 \(\rightarrow\) 1.000 & No corrected signal in either validation mode. \\
PRJEB84421 & OFG vs healthy & 0.019 \(\rightarrow\) 0.019 & 0.282 \(\rightarrow\) 0.282 & Rebuilt support is unchanged because repeated subjects were not detected. \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S11.} Mapping-balance checks for
direct AGP-derived label transfer. Coverage is the percentage of cases or
controls that could be assigned an AGP-derived absorb dCST label at the
specified depth.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} l r r r r Y @{}}
\toprule
Cohort comparison & Depth & Case coverage (\%) & Control coverage (\%) & Fisher p & Interpretation \\
\midrule
Halfvarson IBD vs healthy & 2 & 93.5 & 81.5 & 0.0047 & Cases mapped more often than controls. \\
Halfvarson IBD vs healthy & 4 & 60.3 & 40.7 & 0.0083 & Deeper transfer has case/control mapping imbalance. \\
Halfvarson UC vs healthy & 2 & 96.4 & 81.5 & 2.9e-04 & UC transfer is strong but mapping is phenotype-associated. \\
Halfvarson UC vs healthy & 4 & 67.0 & 40.7 & 3.7e-04 & Deeper UC transfer requires cautious interpretation. \\
HMP2 IBD vs healthy & 2 & 100.0 & 95.6 & 0.072 & High, nearly balanced depth-2 mapping. \\
HMP2 IBD vs healthy & 4 & 52.1 & 66.7 & 0.114 & Deeper mapping drops and reverses direction. \\
Gevers Crohn vs healthy & 2 & 77.7 & 78.8 & 1.000 & Balanced mapping but no transferred association. \\
PRJEB84421 OFG vs healthy & 2 & 79.3 & 85.0 & 0.720 & Balanced mapping but no transferred association. \\
Jacobs IBS vs healthy & 2 & 92.9 & 94.2 & 0.783 & Balanced mapping; IBS transfer remains inconclusive. \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S12.} Absorb-policy robustness for
selected IBD labels. Values compare the absorb dCST view used in the
main manuscript with the pure-policy baseline before low-frequency
dominance patterns are absorbed into retained named states.

\begin{center}
\scriptsize
\begin{tabularx}{\textwidth}{@{} Y r Y Y @{}}
\toprule
Label & Depth & Absorb view & Pure-policy baseline \\
\midrule
\taxon{Bacteroides} & 1 & OR 1.47; q = 3.4e-10 & OR 1.46; q = 1.1e-09 \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae & 2 & OR 3.04; q = 2.9e-28 & OR 3.18; q = 7.0e-30 \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes} & 3 & OR 10.01; q = 5.4e-28 & OR 10.44; q = 7.2e-26 \\
\taxon{Bacteroides}\dcstsep Lachnospiraceae\dcstsep\taxon{Alistipes}\dcstsep\taxon{Faecalibacterium} & 4 & OR 5.54; q = 3.2e-07 & OR 8.15; q = 9.6e-07 \\
\taxon{Morganella} & 1 & OR 16.45; q = 7.7e-17 & OR 18.04; q = 1.8e-17 \\
\taxon{Proteus} & 1 & OR 3.34; q = 0.013 & OR 3.34; q = 0.013 \\
\bottomrule
\end{tabularx}
\end{center}

\bigskip

\noindent\textbf{Supplementary Table S13.} Taxonomic-rank transparency
for selected IBD label components. The dCST labels use the deepest
informative retained-feature annotation, so labels can mix taxonomic
ranks.

\begin{center}
\footnotesize
\begin{tabularx}{\textwidth}{@{} l l Y @{}}
\toprule
Component & Rank used by retained feature & SILVA lineage summary \\
\midrule
\taxon{Bacteroides} & Genus & Bacteroidota; Bacteroidia; Bacteroidales; Bacteroidaceae \\
Lachnospiraceae & Family & Firmicutes; Clostridia; Lachnospirales \\
\taxon{Alistipes} & Genus & Bacteroidota; Bacteroidia; Bacteroidales; Rikenellaceae \\
\taxon{Faecalibacterium} & Genus & Firmicutes; Clostridia; Oscillospirales; Ruminococcaceae \\
\taxon{Morganella} & Genus & Proteobacteria; Gammaproteobacteria; Enterobacterales; Morganellaceae \\
\taxon{Proteus} & Genus & Proteobacteria; Gammaproteobacteria; Enterobacterales; Morganellaceae \\
\bottomrule
\end{tabularx}
\end{center}

\end{document}
