---
title: "Example datasets and reproducible workflows"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Example datasets and reproducible workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style>
.linf-table { max-width: 100%; overflow-x: auto; margin: 1em 0; }
.linf-table:focus { outline: 2px solid #0072b2; outline-offset: 2px; }
.linf-table table { display: table; overflow: visible; width: 100%; table-layout: auto; }
.linf-table table { min-width: 38rem; }
.linf-table td, .linf-table th { vertical-align: top; overflow-wrap: normal; word-break: normal; }
.linf-table td code, .linf-table th code { white-space: nowrap; }
.linf-catalog td, .linf-catalog th { min-width: 12rem; hyphens: none; }
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                     fig.width = 6, fig.height = 4)
library(linf)
# Keep table rendering separate from the runnable analysis code.
knitr::opts_chunk$set(render = function(x, options) {
  rendered <- knitr::knit_print(x, options = options)
  if (inherits(x, "knitr_kable")) {
    knitr::asis_output(paste0(
      '<div class="linf-table" role="region" aria-label="Computed table; scroll horizontally if needed" tabindex="0">\n\n',
      rendered, '\n\n</div>\n'
    ))
  } else rendered
})
```

Choose a matrix if you want to fit a hierarchy, an assignment table if you want
to inspect a saved grouping, and the four-component Valencia subset if you
want to explore coordinate charts. The package supplies five datasets and
small examples can be constructed directly in R; it has no synthetic-data
generation API. The [function guide](function-guide.html) explains how each
stage fits together.

## Choose a dataset

Tables scroll horizontally on narrow screens. Focus a table and use the arrow
keys, or swipe within it, to read every column without breaking function or dataset names.

Every object below is loaded with `data(object_name, package = "linf")` and has
its own help page, for example `help("agp_gut", package = "linf")`.

<div class="linf-table linf-catalog" role="region" aria-label="example-datasets table 1; scroll horizontally if needed" tabindex="0">

| Dataset | Unit, orientation and dimensions | Identifiers and main fields | Suitable workflow |
|---|---|---|---|
| `agp_gut` | 766 gut samples by 314 taxa; integer read-count matrix. | `counts`, `meta`, `taxa`, `source`; row names join to `meta$Run` (SRA accessions). | [Count filtering and fitting](#reference-and-query). |
| `valencia2k` | 2,000 vaginal samples by 178 taxa; unit-sum relative abundances. | `rel`, `cst`, `reads`, `taxa`, `source`; local row IDs join to `cst$sample_id`; `cst$source_row` joins to the pinned source. | [Align metadata](#align-data-and-metadata), then follow the [vaginal tutorial](linf-vaginal.html). |
| `valencia13k_dcst_depth2_merged` | 12,881 source samples; assignment table with six columns, not an abundance matrix. | `assignments`, `summaries`, `feature_labels`, `params`, `source`; `sample_id` and `source_row`. | Inspect saved absorb-policy depth-1/2 assignments and sizes. |
| `valencia13k_dcst_depth3_merged` | Same 12,881 source samples; assignment table with seven columns. | Same fields as depth 2, plus `dcst_depth3` and a depth-3 summary. | Inspect the saved deeper grouping and compare depths. |
| `valencia_linf_hypercube_1k` | 1,000 vaginal samples by four selected taxa; unit-sum matrix. | `rel4`, `meta`, `component_map`, `source`; local `sample_id`, source-row index and selected mass. | [Compute a static chart](#a-static-chart-from-the-four-component-data); coordinates are not precomputed. |

</div>

The two merged objects are ordinary lists of assignments and summary tables.
They lack the complete `linf.csts` object and input matrix required by
`refine.linf.csts()`, `dcst.view()`, `linf.landmarks()` and `transfer.dcsts()`.
Do not change their class to make them look like fresh fits.

```{r inventory}
data(agp_gut, package = "linf")
data(valencia2k, package = "linf")
data(valencia13k_dcst_depth2_merged, package = "linf")
data(valencia13k_dcst_depth3_merged, package = "linf")
data(valencia_linf_hypercube_1k, package = "linf")

observed <- rbind(
  gut_counts = dim(agp_gut$counts),
  vaginal_rel = dim(valencia2k$rel),
  merged_depth2 = dim(valencia13k_dcst_depth2_merged$assignments),
  merged_depth3 = dim(valencia13k_dcst_depth3_merged$assignments),
  four_component_rel = dim(valencia_linf_hypercube_1k$rel4)
)
colnames(observed) <- c("Rows", "Columns")
stopifnot(identical(unname(observed),
                   matrix(c(766L, 2000L, 12881L, 12881L, 1000L,
                            314L, 178L, 6L, 7L, 4L), ncol = 2)))
knitr::kable(observed, caption = "Dimensions read from the bundled objects")
```

## Provenance and scope

### Gut counts

`agp_gut` comes from American Gut Project public 16S V4 records under
[PRJEB11419](https://www.ebi.ac.uk/ena/browser/view/PRJEB11419), processed to
SILVA taxonomic counts through PRIME. The source preparation first selected
up to 5,000 samples with at least 1,000 reads. The 766-sample bundle includes
all samples assigned to four selected uncommon demonstration dCSTs
(Prevotella_7, Pasteurellaceae, Akkermansia and Staphylococcus), excludes
Eukaryota/Unassigned depth-1 labels, and fills remaining slots from the eligible
background using Python's seeded random sampling with seed 42.

The stored `dcst_depth1`/`dcst_depth2` metadata were derived in the source
analysis; refitting on this selected subset need not reproduce them. Phenotype
fields were joined **after** membership was fixed. Selection differs by dCST,
so this is not a probability sample for population inference. Self-reported
phenotypes are available for alignment demonstrations, not general prevalence,
association, causal or clinical claims. They are not independent validation
labels. Counts retain the limitations of 16S taxonomic assignment and the
documented caution about Escherichia-Shigella in V4 data.

### Valencia abundance data and saved assignments

The four Valencia-derived objects use the filtered training source for the
[VALENCIA classifier](https://github.com/ravel-lab/VALENCIA) described by
[France et al. (2020)](https://doi.org/10.1186/s40168-020-00934-6).
The prepared source has 12,881 samples and 178 taxa. `valencia2k` was sampled
with R seed 42 within the 13 supplied sub-CST groups, approximately preserving
their proportions with rounding and a minimum of five per stratum.

`valencia2k$reads` contains post-taxonomic-filter read totals. Multiplying
`rel` by `reads` gives **count-like reconstructions**; rounding them does not
recover or certify original observed integer counts. Use `rel` directly for
normalization and rank-based fitting. The upstream filtering is described in
the existing [vaginal tutorial](linf-vaginal.html); the subset-building script
takes that prepared matrix as input and does not repeat its filtering.

The supplied `Val_CST` and `Val_subCST` labels belong to the Valencia training
source. A comparison made on these bundled rows overlaps that source, even
though dCST fitting does not use its class labels. Such agreement is descriptive,
not held-out validation of a classifier or evidence of clinical performance.

The merged assignment assets were built on the full prepared source with
`n0 = 50`, `low.freq.policy = "absorb"` and `refinement.factor = 2`, with one
or two refinements. Their summaries contain `depth`, `dcst_label`, `n`, `prop`
and `path_length`. Proportions refer to this source sample set. Path lengths
may be shorter than the stored depth for terminal lineages; treat saved labels
as labels, not as a source of stable feature IDs.

### Four-component data

`valencia_linf_hypercube_1k` selects Lactobacillus iners (`Li`), Lactobacillus
crispatus (`Lc`), Gardnerella vaginalis (`Gv`) and BVAB1 (`Bv`). Rows with zero
total mass in those four taxa were removed, and retained rows were renormalized
to sum to one **over those four features**. R seed 20261604 was used for a
1,000-row sample stratified by the dominant selected component.

`meta$selected_mass` records how much of the original full profile these four
taxa represented. A large fraction within `rel4` can still correspond to little
of the full community. `component_map` connects abbreviations to source taxa;
neither the subset nor its charts replace the full 178-feature profiles.

All objects include a provenance string. The installed provenance file records
source preparation and upstream terms (which are separate from the code's MIT
license):

```{r provenance-location}
provenance <- system.file("DATA_PROVENANCE.md", package = "linf", mustWork = TRUE)
cat(readLines(provenance, warn = FALSE)[1:12], sep = "\n")
```

Read the full installed record at the console with
`cat(readLines(provenance), sep = "\n")`. This works offline. Repository readers
can open [data provenance](https://github.com/pgajer/linf/blob/main/inst/DATA_PROVENANCE.md),
the [input manifest](https://github.com/pgajer/linf/blob/main/inst/DATA_MANIFEST.csv),
and the [reproduction recipe](https://github.com/pgajer/linf/blob/main/data-raw/README.md).

## Align data and metadata

Sample names are keys within an object, not universal cross-dataset IDs. For
`valencia2k`, IDs such as `s1` were reassigned after sampling. Its verified
`cst$source_row` now preserves the sampled row in the pinned prepared source.
The four-component and merged objects share that `source_row` numbering.
Join on this explicit key after checking provenance, never by row position or
an apparent numeric suffix in the different local sample-ID schemes.

```{r alignment}
X <- agp_gut$counts
stopifnot(!anyDuplicated(agp_gut$meta$Run),
          !anyDuplicated(agp_gut$taxa))
sample.order <- match(rownames(X), agp_gut$meta$Run)
feature.order <- match(colnames(X), agp_gut$taxa)
stopifnot(!anyNA(sample.order), !anyNA(feature.order))
meta <- agp_gut$meta[sample.order, , drop = FALSE]
features <- data.frame(
  id = paste0("asv_", seq_len(ncol(X))),
  taxonomy = agp_gut$taxa[feature.order]
)
stopifnot(identical(rownames(X), meta$Run))

# After filtering, indices refer to the supplied X, in its original order.
filtered <- filter.asv(X, min.lib = 1000, prev.prop = 0.05,
                       min.count = 2, verbose = FALSE)
meta.filtered <- meta[filtered$kept.sample.idx, , drop = FALSE]
features.filtered <- features[filtered$kept.feature.idx, , drop = FALSE]
stopifnot(identical(rownames(filtered$counts), meta.filtered$Run),
          nrow(features.filtered) == ncol(filtered$counts))
```

The stable IDs here describe the original column order and are subsetted with
the features; they are not regenerated after filtering. Pass
`features.filtered$id` and display labels made from its taxonomy in that same
order when fitting. Keep the mapping with the fit. Landmarks preserve complete stable IDs, including taxonomy strings containing
`__`; see [landmark interpretation](function-guide.html#landmarks-are-selected-observations).

The following join uses the explicit shared source-row key:

```{r shared-source-join}
merged <- valencia13k_dcst_depth3_merged
chart.meta <- valencia_linf_hypercube_1k$meta
stopifnot(!anyDuplicated(merged$assignments$source_row))
join <- match(chart.meta$source_row, merged$assignments$source_row)
stopifnot(!anyNA(join))
chart.depth3 <- merged$assignments$dcst_depth3[join]
stopifnot(identical(as.character(chart.meta$Val_CST),
                   as.character(merged$assignments$Val_CST[join])))
length(chart.depth3)
subset.join <- match(valencia2k$cst$source_row, merged$assignments$source_row)
stopifnot(!anyNA(subset.join),
          identical(as.character(valencia2k$cst$Val_CST),
                    as.character(merged$assignments$Val_CST[subset.join])))
```

This attaches in-source saved labels; it is not a new reference/query
experiment.

## Constructed edge cases

This deterministic matrix needs no random seed. It includes two supported
dominant features, a low-support feature, a positive tie, an exact-zero row and
a small row left unchanged by a positive tolerance. IDs differ from labels.

```{r toy-edge-cases}
toy <- rbind(
  a1 = c(10, 3, 1), a2 = c(9, 2, 1),
  b1 = c(2, 10, 1), b2 = c(3, 9, 1),
  low_support = c(1, 2, 10), tie = c(4, 4, 0),
  zero = c(0, 0, 0), small = c(0.01, 0.005, 0)
)
colnames(toy) <- c("asv_1", "asv_2", "asv_3")
display <- c("Taxon A", "Taxon B", "Taxon C")
M <- normalize.linf(toy, tol = 0.02)
d <- linf.dominant.features(M, feature.labels = display)
fit <- linf.csts(M, feature.labels = display, n0 = 2)
edge.table <- data.frame(
  sample = rownames(M), maximum = apply(M, 1, max),
  dominant_ID = d$id, pure = fit$lineage.label.pure,
  absorb = fit$lineage.label.absorb
)
knitr::kable(edge.table, row.names = FALSE)
stopifnot(is.na(d$id[7]), d$id[6] == "asv_1", d$id[8] == "asv_1",
          identical(unname(M["small", ]), unname(toy["small", ])))
```

The tied row chooses the first column. The small row retains maximum 0.01
and still has a dominant feature. The zero row has no dominant feature; pure
fitting puts it in the rare bucket and absorb fitting uses the most-supported
retained state. The low-support row is absorbed into Taxon B, its largest
retained feature. Neither absorption is evidence that the assigned feature was
the original maximum.

For random ties, reset the seed before each call you want to reproduce:

```{r random-tie}
set.seed(17)
random1 <- linf.dominant.features(M, tie.method = "random")$id
set.seed(17)
random2 <- linf.dominant.features(M, tie.method = "random")$id
stopifnot(identical(random1, random2))
```

## Reference and query

Use a small seeded sample of 160 Valencia rows. The first 120 sampled rows form
the reference; the other 40 are kept out of fitting and refinement. This split
demonstrates frozen transfer mechanics within one source dataset, not external
validation, and no outcome or CST label is used to fit the reference.

```{r real-transfer}
set.seed(20260914)
rows <- sample(seq_len(nrow(valencia2k$rel)), 160)
reference.rows <- rows[1:120]
query.rows <- rows[121:160]
stopifnot(length(intersect(reference.rows, query.rows)) == 0L)
reference <- valencia2k$rel[reference.rows, , drop = FALSE]
query <- valencia2k$rel[query.rows, , drop = FALSE]
reference.M <- normalize.linf(reference)
reference.fit <- linf.csts(reference.M, n0 = 5,
                           low.freq.policy = "absorb")
reference.fit <- refine.linf.csts(
  reference.M, reference.fit, n0 = 3, refinement.factor = 2,
  low.freq.policy = "absorb", verbose = FALSE
)

# Deliberately reverse query columns. Matching uses the unchanged feature IDs.
query <- query[, rev(seq_len(ncol(query))), drop = FALSE]
stopifnot(setequal(colnames(query), reference.fit$feature.ids))
transferred <- transfer.dcsts(query, reference.fit, match.by = "feature.ids")
knitr::kable(data.frame(
  depth = colnames(transferred$assignment),
  assigned = colSums(!is.na(transferred$assignment)),
  unassigned = colSums(is.na(transferred$assignment))
), row.names = FALSE)
```

No query filtering rule or hierarchy is learned here. Missing reference columns
would be filled with zero; extra query columns would be ignored. Counts and
positive row-scaled abundances give the same ranking for this transfer.
`NA` means the walk found no positive realized candidate at that depth; it does
not create a new dCST. A partial depth-1 assignment can coexist with a depth-2
`NA`.

```{r no-matching-feature}
unmatched <- matrix(5, nrow = 1, ncol = 1,
                    dimnames = list("unmatched_query", "not_in_reference"))
no.assignment <- transfer.dcsts(unmatched, reference.fit)$assignment
stopifnot(all(is.na(no.assignment)))
no.assignment
```

For a longer count-based workflow use the [gut demonstration](linf-intro.html);
for descriptive comparisons to source CST labels use the
[vaginal workflow](linf-vaginal.html).

## A static chart from the four-component data

The returned chart has three coordinates. This plot shows two of them for all
1,000 bundled samples; color identifies which of the four selected components
has the largest abundance. It is a projection of a representation, with no
learned classifier or biological-separation test.

```{r static-chart, fig.cap="Two coordinates of the Li-reference chart. Colors describe dominance within the selected four taxa; overlapping points and the omitted third coordinate limit what this projection shows."}
four <- valencia_linf_hypercube_1k
embedding <- linf.hypercube.embedding(four$rel4, reference = "Li")
stopifnot(identical(dim(embedding), c(1000L, 3L)),
          all(is.finite(embedding)))
colors <- c(Li = "#0072B2", Lc = "#D55E00", Gv = "#009E73", Bv = "#CC79A7")
plot(embedding[, 1], embedding[, 2],
     col = colors[as.character(four$meta$dominant_component)],
     pch = 16, cex = 0.5, xlim = c(0, 1), ylim = c(0, 1),
     xlab = "Lc coordinate (reference Li)",
     ylab = "Gv coordinate (reference Li)")
legend("topright", legend = names(colors), col = colors, pch = 16,
       title = "Dominant component", bty = "n", cex = 0.8)
```

Check `summary(four$meta$selected_mass)` before interpreting the four-taxon
composition. The optional [interactive embedding article](https://pgajer.github.io/linf/articles/valencia-hypercube-embedding.html)
adds three-dimensional views and is website-only. The four installed vignettes
are listed by `browseVignettes("linf")`; their normal builds need no network
downloads, interactive device or external dataset.
