---
title: "DCSTs for Vaginal Microbiome Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DCSTs for Vaginal Microbiome Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(linf)
```

## Introduction

Community state types (CSTs) are the standard way to describe the structure of
vaginal microbial communities. They were originally defined by hierarchical
clustering of 16S rRNA data (Ravel et al. 2011) and later standardized by the
VALENCIA nearest-centroid classifier (France et al. 2020).

**Dominant Community State Types (DCSTs)** offer an alternative that requires no
clustering and no training data. A DCST is defined by
$L^\infty$-normalization: divide each sample's abundance vector by its maximum
value, then group samples by which taxon achieves that maximum. This is a
purely geometric operation on compositional data --- it identifies the dominant
species in each sample and groups samples accordingly.

This vignette demonstrates the DCST workflow on a real vaginal 16S dataset and
shows that DCSTs recover the established CST structure with high concordance.

**Reference:** Gajer, P. & Ravel, J. (2025). A New Approach to Compositional
Data Analysis using $L^\infty$-normalization with Applications to Vaginal
Microbiome. arXiv:2503.21543.

## The Valencia 2k dataset

The `valencia2k` dataset is a stratified subsample of 2,000 vaginal samples from
the Valencia CST-classifier training set (France et al. 2020). It contains
species-level relative abundances for 178 taxa along with the original Valencia
CST and sub-CST assignments.

```{r load-data}
data(valencia2k)

## Compositional matrix: 2000 samples x 178 taxa
dim(valencia2k$rel)

## Valencia CST assignments
table(valencia2k$cst$Val_CST)

## Top 10 most abundant taxa (by mean relative abundance)
means <- sort(colMeans(valencia2k$rel), decreasing = TRUE)
head(round(means, 4), 10)
```

The dominant taxa are familiar vaginal species: *Lactobacillus iners*,
*L. crispatus*, *Gardnerella vaginalis*, and others. This is exactly the kind
of data where $L^\infty$-cells are most intuitive: most samples are dominated
by a single species, and the identity of that species is biologically meaningful.

## Reconstructing counts

The bundled matrix is compositional (rows sum to 1). For workflows that require
count-like input (e.g., `filter.asv()`), we can reconstruct approximate counts
using the per-sample read depths:

```{r reconstruct-counts}
count_mat <- sweep(valencia2k$rel, 1, valencia2k$reads, "*")
count_mat <- round(count_mat)
storage.mode(count_mat) <- "integer"

## Check: library sizes
summary(rowSums(count_mat))
```

Since the Valencia 13k data was already filtered (min 3,000 reads, taxa present
in $\ge$ 1% of samples), we skip `filter.asv()` here and work directly with
the compositional matrix.

## $L^\infty$ normalization

$L^\infty$-normalization divides each row by its maximum value. After
normalization, every sample has at least one coordinate equal to 1, and all
other coordinates lie in $[0, 1]$.

```{r linf-normalize}
Z <- normalize.linf(valencia2k$rel)

## Every row max should be 1
summary(apply(Z, 1, max))
```

## $L^\infty$-cells: dominant species assignment

An $L^\infty$-cell is simply the column that achieves the maximum in each row.
This is a deterministic, sample-independent assignment: a sample's cell
membership depends only on its own abundance profile, never on other samples.

```{r linf-cells}
cells <- linf.cells(Z)

## How many distinct dominant taxa?
cat("Distinct L-inf cells:", length(cells$observed.levels), "\n")

## Cell size distribution (top 15)
cell_tab <- sort(table(cells$label), decreasing = TRUE)
head(cell_tab, 15)
```

We see the expected vaginal community structure: a few large cells (the
*Lactobacillus* species, *Gardnerella*, BVAB1) and a long tail of rarer
dominant taxa.

## Depth-1 DCSTs

In practice, many $L^\infty$-cells contain very few samples. **Truncated
$L^\infty$-CSTs (DCSTs)** keep only cells above a minimum size threshold
($n_0$) and reassign small-cell samples to the nearest large cell.

The package supports two policies for handling small cells:

- **"pure"**: named DCSTs stay pure, and small-cell samples are labeled `RARE_DOMINANT`
- **"absorb"**: small-cell samples are reassigned to the nearest kept cell by
  restricted argmax

The legacy policy name `"rare"` is still accepted as a deprecated alias for `"pure"`.

```{r depth1-dcsts}
dcst1 <- linf.csts(Z, n0 = 30, low.freq.policy = "pure")

## Kept cells
dcst1$kept.cells.lbl

## DCST assignment (pure policy)
dcst1_tab <- sort(table(dcst1$cell.label), decreasing = TRUE)
dcst1_tab
```

### Concordance with Valencia CSTs

The key test: do DCSTs correspond to the established Valencia CSTs? Let's
build a cross-tabulation:

```{r concordance-depth1}
## Build concordance table: rows = DCSTs, columns = Valencia CSTs
dcst_labels <- dcst1$cell.label.absorb  # use absorb view for clean comparison
val_cst     <- valencia2k$cst$Val_CST

concordance <- table(DCST = dcst_labels, Valencia = val_cst)

## Show as percentage: what fraction of each DCST falls in each Valencia CST?
pct <- round(100 * prop.table(concordance, margin = 1), 1)
pct
```

The concordance is striking. Each DCST maps predominantly to a single Valencia
CST:

- *Lactobacillus crispatus* DCST $\to$ CST I (dominated by *L. crispatus*)
- *Lactobacillus iners* DCST $\to$ CST III (dominated by *L. iners*)
- *Gardnerella vaginalis* DCST $\to$ CST IV-B
- BVAB1 DCST $\to$ CST IV-A
- *Lactobacillus gasseri* DCST $\to$ CST II
- *Lactobacillus jensenii* DCST $\to$ CST V

```{r concordance-reverse}
## Reverse view: what fraction of each Valencia CST falls in each DCST?
pct_rev <- round(100 * prop.table(concordance, margin = 2), 1)
pct_rev
```

This two-way concordance confirms that DCSTs are an alternative construct of
CSTs: they recover the same groupings without any clustering algorithm or
training data.

## Depth-2 refinement

Large DCSTs can be refined by dropping the dominant species and repeating the
procedure on the remaining taxa. This produces a two-level hierarchy that
reveals sub-community structure.

```{r depth2-refine}
dcst2 <- refine.linf.csts(Z, dcst1, n0 = 15, refinement.factor = 2,
                           low.freq.policy = "pure", verbose = FALSE)

## Depth-2 DCST table
dcst2_tab <- sort(table(dcst2$cell.label), decreasing = TRUE)
dcst2_tab
```

### Concordance with Valencia sub-CSTs

```{r concordance-depth2}
dcst2_labels <- dcst2$cell.label.absorb
val_subcst   <- valencia2k$cst$Val_subCST

concordance2 <- table(DCST = dcst2_labels, Valencia = val_subcst)

## Percentage by DCST (row-wise)
pct2 <- round(100 * prop.table(concordance2, margin = 1), 1)

## Show only DCSTs with at least 20 samples for readability
dcst2_sizes <- rowSums(concordance2)
pct2[dcst2_sizes >= 20, , drop = FALSE]
```

The depth-2 refinement recovers sub-CST structure. For example, the
*Lactobacillus crispatus* DCST splits based on the second-most-abundant taxon,
and these sub-DCSTs align with Valencia sub-CSTs I-A and I-B.

## The full pipeline

For convenience, `linf.dcst.landmark.pipeline()` runs the entire workflow in
one call: normalization, depth-1 DCSTs, depth-2 refinement, and landmark
computation.

```{r full-pipeline}
out <- linf.dcst.landmark.pipeline(
  count_mat,
  feature.ids    = colnames(count_mat),
  feature.labels = colnames(count_mat),
  n0.depth1      = 30,
  n0.depth2      = 15,
  refinement.factor = 2,
  low.freq.policy   = "pure",
  landmark.view     = "absorb",
  verbose           = FALSE
)

names(out)
```

## Landmark points

Each DCST cell has characteristic landmark points: the most extreme sample
(highest value of the dominant taxon), the least extreme (lowest value while
still dominant), and mean/median representatives. These help characterize what
a "typical" or "extreme" member of each community type looks like.

```{r landmarks}
## Depth-1 landmarks
lm1 <- out$landmarks.depth1$landmarks

## Show landmarks for a few key DCSTs
key_cells <- c("Lactobacillus_crispatus", "Lactobacillus_iners",
               "Gardnerella_vaginalis")
key_lm <- lm1[lm1$cell.id %in% key_cells, ]

## For each DCST, show the endpoint.max landmark's target value
## (how dominant is the dominant species in the most extreme sample?)
ep_max <- key_lm[key_lm$landmark.type == "endpoint.max",
                 c("cell.label", "point.name", "target.value")]
ep_max

## And the endpoint.min (least dominant while still assigned to this cell)
ep_min <- key_lm[key_lm$landmark.type == "endpoint.min",
                 c("cell.label", "point.name", "target.value")]
ep_min
```

The endpoint.max landmarks show samples where the dominant species comprises
nearly the entire community, while endpoint.min landmarks show the boundary
cases --- samples where dominance is marginal.

## Summary

This vignette demonstrated the DCST workflow on real vaginal 16S data:

1. **$L^\infty$-normalization** maps compositional data to the $L^\infty$-simplex
2. **$L^\infty$-cells** assign each sample to its dominant taxon
3. **Truncated DCSTs** keep frequent cells and handle rare dominance patterns
4. **Depth-2 refinement** reveals sub-community structure
5. **Landmark points** characterize extreme and typical community members

The concordance with established Valencia CSTs is high, validating DCSTs as a
principled, deterministic alternative to clustering-based community typing. Key
advantages:

- **No clustering algorithm:** assignment depends only on the sample itself
- **No training data:** new samples can be assigned without a reference database
- **Interpretable:** DCST labels directly name the dominant species
- **Hierarchical:** refinement naturally produces sub-types

## References

- France, M. T., Ma, B., Gajer, P., et al. (2020). VALENCIA: a nearest
  centroid classification method for vaginal microbial communities based on
  composition. *Microbiome*, 8(1), 166.
- Gajer, P. & Ravel, J. (2025). A New Approach to Compositional Data Analysis
  using $L^\infty$-normalization with Applications to Vaginal Microbiome.
  arXiv:2503.21543.
- Ravel, J., Gajer, P., Abdo, Z., et al. (2011). Vaginal microbiome of
  reproductive-age women. *PNAS*, 108(Suppl 1), 4680--4687.
