---
title: "Dominant Community State Types: From Normalization to a Gut Demonstration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dominant Community State Types: From Normalization to a Gut Demonstration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
table td, table th { white-space: normal; overflow-wrap: anywhere; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
library(linf)
```

The `linf` package implements L-infinity normalization and Dominant Community
State Types (dCSTs) for compositional data. This vignette demonstrates the
complete workflow in two parts: first a quick-start with toy data to introduce
the core functions, then a real-data analysis of gut microbiome samples from
the American Gut Project showing how to inspect dominant and subdominant
community structure.

For entry points and policy contracts see the [function guide](function-guide.html);
for dataset selection and safe joins see [example datasets](example-datasets.html).

## Part 1: Quick Start with Toy Data

### L-infinity normalization

L-infinity normalization scales each row of a count matrix by its maximum
value when that maximum exceeds `tol`. At the default `tol = 0`, positive
rows have maximum one and exact-zero rows remain zero. With positive tolerance,
rows at or below it remain unchanged, including small nonzero entries.

```{r toy-normalize}
set.seed(1)
S.counts <- matrix(
  rpois(30, lambda = 5),
  nrow = 10, ncol = 3,
  dimnames = list(paste0("s", 1:10), c("Taxon_A", "Taxon_B", "Taxon_C"))
)

Z <- normalize.linf(S.counts)
apply(Z, 1, max)   # all 1
```

### Dominance sample sets

Each sample is assigned to its dominant feature: the column achieving the
within-sample maximum. Samples with the same dominant feature form a
depth-1 *dominance sample set*.

```{r toy-dominant.features}
dominant.features <- linf.dominant.features(Z)
table(dominant.features$label)
```

### Truncated dCSTs

`linf.csts()` groups samples by dominant feature and collapses small groups
(below threshold `n0`) into a `RARE_DOMINANT` bucket. This produces
*Dominant Community State Types* (dCSTs).

```{r toy-csts}
A <- matrix(c(5, 1, 0), nrow = 6, ncol = 3, byrow = TRUE)
B <- matrix(c(1, 5, 0), nrow = 2, ncol = 3, byrow = TRUE)
C <- matrix(c(1, 0, 5), nrow = 2, ncol = 3, byrow = TRUE)
S <- rbind(A, B, C)
S <- sweep(S, 1, rowSums(S), "/")
colnames(S) <- c("Dom1", "Dom2", "Dom3")

res <- linf.csts(S, n0 = 5)
table(res$lineage.label, useNA = "ifany")
```

Only Dom1 (6 samples) meets the `n0 = 5` threshold. Dom2 and Dom3 are
collapsed into `RARE_DOMINANT`.


## Part 2: Gut Microbiome dCST Demonstration

### Background

Single-species dominance is common in some microbial ecosystems and less
common in others. With default first-column tie handling, dCSTs provide a deterministic way to summarize the dominant
feature and then refine large groups by subdominant features. The gut example
below demonstrates those mechanics without treating the selected package
subset as an epidemiologic sample.

### The data

The `agp_gut` dataset bundled with this package contains `r nrow(agp_gut$counts)`
gut microbiome samples from the American Gut Project (PRJEB11419), a citizen-science
16S rRNA V4 survey. The dataset is a dCST-stratified demonstration subset
that includes all samples in four selected uncommon dCSTs and fills
the remaining slots with a seed-42 simple random sample from the eligible
background. Phenotypes are joined only after selection and do not influence
membership. Because inclusion probabilities differ by dCST, it is not a
probability sample of the underlying cohort; phenotype frequencies, effect
sizes, and hypothesis tests must not be interpreted as population results.

```{r load-data}
data(agp_gut)
str(agp_gut, max.level = 1)

dim(agp_gut$counts)   # samples x taxa
```

Metadata includes self-reported health conditions parsed from the AGP
questionnaire:

```{r meta-summary}
disease_cols <- c("IBS", "IBD", "Obesity", "Cardiovascular_disease",
                  "Autoimmune", "Acid_reflux")
sapply(disease_cols, function(col) sum(agp_gut$meta[[col]], na.rm = TRUE))
```

### Step 1: Quality filtering

`filter.asv()` removes low-depth samples and rare taxa in one call. We
require at least 1,000 reads per sample and taxa present in at least 5% of
samples.

```{r filter}
filt <- filter.asv(agp_gut$counts, min.lib = 1000, prev.prop = 0.05,
                   min.count = 2)
dim(filt$counts)
dim(filt$rel)
```

### Step 2: L-infinity normalization

```{r normalize}
M <- normalize.linf(filt$counts)

# Verify: every row max is 1
stopifnot(all(abs(apply(M, 1, max) - 1) < 1e-10))
```

### Step 3: Depth-1 dCSTs

```{r dcst-depth1}
csts <- linf.csts(M, n0 = 30)

# dCST size distribution
sort(table(csts$lineage.label), decreasing = TRUE)
```

The largest retained groups are *Bacteroides* and *Escherichia-Shigella*.
The figure below labels all seven retained groups and the separate rare bucket.
This tutorial uses the pure policy; the README figure uses absorption, so its
group counts differ even though both start with the same filtered samples.
The `RARE_DOMINANT` bucket collects samples dominated by taxa too infrequent
to form their own dCST at this threshold.

The *Escherichia-Shigella* values are processed 16S read counts.
[McDonald et al. (2018)](https://doi.org/10.1128/mSystems.00031-18) describe
microbial overgrowth during room-temperature shipping in American Gut,
including *E. coli* blooms. That source does not establish the cause of the
abundance in this bundled subset. The displayed dominance count alone cannot
establish organismal abundance or a particular technical cause, including
primer cross-reactivity.

### How sensitive is depth 1 to the support threshold?

The following comparison uses the same `r nrow(M)` filtered samples and
`r ncol(M)` features at every setting. Only the minimum support and the
low-support policy change; no phenotype information enters these fits.

```{r support-sensitivity}
sensitivity <- do.call(rbind, lapply(c(10L, 30L, 60L, 200L), function(support) {
  do.call(rbind, lapply(c("pure", "absorb"), function(policy) {
    fit <- linf.csts(M, n0 = support, low.freq.policy = policy,
                     return.diagnostics = TRUE)
    counts <- summary(fit)
    stopifnot(counts$input.samples == nrow(M),
              counts$input.samples == counts$assigned.samples + counts$unassigned.samples)
    data.frame(n0 = support, Policy = policy, Groups = counts$retained.lineages,
               Rare = counts$rare.samples,
               Reassigned = if (policy == "absorb") sum(fit$diagnostics$reassigned) else 0L,
               Unassigned = counts$unassigned.samples)
  }))
}))
knitr::kable(sensitivity)
```

At `n0 = 30`, the pure view has
`r sensitivity$Rare[sensitivity$n0 == 30 & sensitivity$Policy == "pure"]`
samples in a synthetic rare category; absorption reassigns
`r sensitivity$Reassigned[sensitivity$n0 == 30 & sensitivity$Policy == "absorb"]`
samples to the retained groups. Raising the support threshold reduces the
number of retained groups. At 200, no feature has enough provisional support:
the pure view puts every sample in the rare category, while absorption leaves
every sample unassigned. Rare samples are assigned to a synthetic category;
"Unassigned" means an `NA` assignment and is counted separately.

This is a descriptive sensitivity check, not a search for an optimal or
clinically validated threshold. Stable group counts would not by themselves
establish stable membership, biological validity or predictive accuracy. The
stratified sampling limitation applies to every row of this comparison.

### Step 4: Depth-2 refinement

`refine.linf.csts()` subdivides large dCSTs by their *subdominant* species.
For instance, a Bacteroides-dominated sample might be further classified by
whether its second-most-abundant taxon is *Faecalibacterium* or
*Lachnospiraceae*. This captures co-dominance patterns that are especially
important in the diverse gut environment.

```{r dcst-depth2}
csts2 <- refine.linf.csts(M, csts, n0 = 30)

# Show depth-2 dCSTs with >= 20 samples
tab2 <- sort(table(csts2$lineage.label), decreasing = TRUE)
tab2[tab2 >= 20]
```

### Dominance strength

A useful diagnostic: the relative abundance of the dominant species in each
sample. In the gut, most samples have modest dominance (< 50%), consistent
with the diverse nature of the ecosystem.

```{r dominance-dist, fig.cap = "Distribution of dominance strength across gut samples. Most samples show moderate dominance, with a tail of strongly dominated communities."}
rel <- filt$rel
dom_strength <- apply(rel, 1, max)
hist(dom_strength, breaks = 50, col = "#2ecc71", border = "white",
     main = "Dominance Strength in Gut Microbiome",
     xlab = "Relative abundance of dominant species",
     ylab = "Number of samples")
abline(v = 0.5, col = "red", lty = 2, lwd = 2)
legend("topright", "50% dominance", col = "red", lty = 2, lwd = 2, bty = "n")
```


### dCST size distribution

```{r dcst-barplot, fig.cap = "Pure-policy depth-1 assignments for 763 filtered AGP samples: seven retained groups and 159 samples in the synthetic rare bucket. Counts describe the stratified demonstration subset, not population frequencies."}
tab1 <- sort(table(csts$lineage.label), decreasing = TRUE)
# Use the last available taxonomy rank for display; keep full IDs in the fit.
short_taxon <- function(id) {
  if (id == "RARE_DOMINANT") return("Rare dominant")
  ranks <- sub("^[a-z]__", "", strsplit(id, ";", fixed = TRUE)[[1L]])
  ranks <- ranks[nzchar(ranks) & ranks != "__"]
  tail(ranks, 1L)
}
names(tab1) <- make.unique(vapply(names(tab1), short_taxon, character(1)))
par(mar = c(5, 10, 3, 2))
bp <- barplot(rev(tab1), horiz = TRUE,
              col = ifelse(rev(names(tab1)) == "Rare dominant", "#e74c3c", "#3498db"),
              las = 1, cex.names = 0.8, xlim = c(0, max(tab1) * 1.18),
              xlab = "Number of samples", main = "Gut dCST sizes (pure policy, depth 1)")
text(rev(tab1), bp, labels = rev(tab1), pos = 4, cex = 0.8)
```


### Phenotype metadata

The bundle retains selected self-reported phenotype fields so that users can
inspect the object structure and practice data alignment. The following table
is purely descriptive; dCST-stratified selection precludes population
prevalence estimation or treating association calculations as cohort results.

```{r phenotype-summary}
meta <- agp_gut$meta[match(rownames(M), agp_gut$meta$Run), ]
stopifnot(all(meta$Run == rownames(M)))

conditions <- c("IBS", "IBD", "Obesity", "Cardiovascular_disease",
                "Autoimmune", "Acid_reflux", "Lung_disease")
phenotype_summary <- data.frame(
  Condition = conditions,
  Recorded_cases = vapply(
    conditions,
    function(x) sum(meta[[x]] == 1, na.rm = TRUE),
    integer(1)
  ),
  Non_missing = vapply(
    conditions,
    function(x) sum(!is.na(meta[[x]])),
    integer(1)
  )
)
knitr::kable(
  phenotype_summary,
  caption = "Recorded phenotypes in the selected demonstration subset"
)
```


## Limitations

This analysis has several important caveats:

- **Self-reported metadata.** AGP health conditions are self-reported by
  citizen-science participants, with unknown accuracy.
- **16S resolution.** The Escherichia-Shigella inflated abundance is a known
  V4 artifact. Shotgun metagenomics would resolve this.
- **dCST-stratified demonstration subset.** Four uncommon dCSTs are included
  exhaustively and the eligible background is sampled with seed 42.
  Phenotypes do not affect membership, but differing inclusion probabilities
  mean the object must not be used for population prevalence estimates,
  effect-size estimation, or association testing.

## Summary

This workflow summarizes dominant features and refines selected groups using
remaining features. The support threshold, refinement selection, low-support
policy and tie handling determine the stored hierarchy. Its descriptive
patterns apply to this selected dataset; they do not establish population
associations or clinical utility.
