---
title: "Linf: L-Infinity Normalization and Dominant Community State Types in R"
date: "2026-08-20"
abstract: >
  The linf package provides dependency-light tools for L-infinity
  normalization of nonnegative compositional matrices and for constructing
  dominant Community State Types (dCSTs). A sample is first assigned to its
  largest feature. Well-supported assignments define retained states, while
  an absorb policy reassigns samples from smaller states among the retained
  features. The same construction can be refined by successive subdominant
  features and transferred to new matrices through a frozen hierarchy. The
  package separates stable feature identifiers from display labels, supports
  dense and sparse matrices, preserves observed zeros, and includes
  zero-aware hypercube coordinates and representative landmarks. We describe
  the software design and demonstrate it with bundled vaginal and gut
  microbiome data. In the vaginal example, dCSTs correspond closely, but not
  identically, to established VALENCIA Community State Types. The package is
  intended as an interpretable compositional summary and does not replace
  log-ratio inference or establish that microbial communities form discrete
  clusters.
draft: true
author:
  - name: Pawel Gajer
    affiliation: University of Maryland School of Medicine
    address:
      - Center for Advanced Microbiome Research and Innovation
      - Institute for Genome Sciences
      - Baltimore, Maryland, USA
    email: pgajer@gmail.com
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble: |
  \usepackage{float}
packages:
  cran:
    - linf
    - Matrix
    - compositions
    - robCompositions
    - zCompositions
  bioc: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  fig.align = "center",
  out.width = "100%",
  dev = if (knitr::is_latex_output()) "cairo_pdf" else "png",
  dpi = 150
)

library(linf)
library(Matrix)

stopifnot(packageVersion("linf") >= "0.2.0")

pretty_taxon <- function(x) {
  x <- sub("^g_", "", x)
  gsub("_", " ", x, fixed = TRUE)
}

short_silva_label <- function(x) {
  ranks <- strsplit(x, ";", fixed = TRUE)[[1L]]
  labels <- sub("^[a-z]__", "", ranks)
  labels <- labels[nzchar(labels) & labels != "__"]
  if (!length(labels)) x else labels[[length(labels)]]
}

output_alt <- function(text) {
  if (knitr::is_html_output()) text else NULL
}

data(valencia2k, package = "linf")
valencia_linf <- normalize.linf(valencia2k$rel)
valencia_dominant <- linf.dominant.features(valencia_linf)
valencia_fit <- linf.csts(
  valencia_linf,
  n0 = 30,
  low.freq.policy = "absorb"
)
valencia_fit_depth2 <- refine.linf.csts(
  valencia_linf,
  valencia_fit,
  n0 = 15,
  refinement.factor = 2,
  low.freq.policy = "absorb",
  verbose = FALSE
)
valencia_concordance <- table(
  dCST = valencia_fit$lineage.label,
  VALENCIA = valencia2k$cst$Val_CST
)
valencia_concordance_pct <- 100 * prop.table(valencia_concordance, margin = 1)
valencia_majority_fraction <-
  sum(apply(valencia_concordance, 1, max)) / sum(valencia_concordance)

set.seed(20260820)
training_index <- sort(sample(seq_len(nrow(valencia_linf)), 1500))
test_index <- setdiff(seq_len(nrow(valencia_linf)), training_index)
transfer_fit <- linf.csts(
  valencia_linf[training_index, ],
  n0 = 25,
  low.freq.policy = "absorb"
)
transfer_fit <- refine.linf.csts(
  valencia_linf[training_index, ],
  transfer_fit,
  n0 = 12,
  refinement.factor = 2,
  low.freq.policy = "absorb",
  verbose = FALSE
)
transferred <- transfer.dcsts(
  valencia_linf[test_index, ],
  transfer_fit,
  depth = 1:2,
  view = "absorb"
)
transfer_concordance <- table(
  dCST = transferred$assignment[, "depth1"],
  VALENCIA = valencia2k$cst$Val_CST[test_index]
)
transfer_majority_fraction <-
  sum(apply(transfer_concordance, 1, max)) / sum(transfer_concordance)

data(agp_gut, package = "linf")
gut_filtered <- suppressMessages(filter.asv(
  agp_gut$counts,
  min.lib = 1000,
  prev.prop = 0.05,
  min.count = 2
))
gut_linf <- normalize.linf(gut_filtered$counts)
gut_fit <- linf.csts(
  gut_linf,
  n0 = 30,
  low.freq.policy = "absorb"
)
gut_sizes <- sort(table(gut_fit$lineage.label), decreasing = TRUE)
names(gut_sizes) <- make.unique(vapply(
  names(gut_sizes), short_silva_label, character(1L)
))
gut_dominance <- apply(gut_filtered$rel, 1, max)

make_sparse_example <- function(n, p = 1000, density = 0.01, seed = 1L) {
  set.seed(seed + n)
  x <- Matrix::rsparsematrix(
    n,
    p,
    density = density,
    rand.x = function(k) rexp(k)
  )
  x@x <- abs(x@x)
  zero_rows <- which(Matrix::rowSums(x) == 0)
  if (length(zero_rows)) {
    x[cbind(zero_rows, ((zero_rows - 1L) %% p) + 1L)] <- 1
  }
  x
}

benchmark_linf <- function(n, p = 1000, repetitions = 5L) {
  sparse_x <- make_sparse_example(n, p)
  dense_x <- as.matrix(sparse_x)
  run <- function(x, backend) {
    median(replicate(repetitions, system.time({
      z <- normalize.linf(x, backend = backend)
      linf.csts(
        z,
        n0 = 10,
        low.freq.policy = "absorb",
        backend = backend
      )
    })[["elapsed"]]))
  }
  data.frame(
    Samples = n,
    Features = p,
    `Dense MB` = as.numeric(object.size(dense_x)) / 1024^2,
    `Sparse MB` = as.numeric(object.size(sparse_x)) / 1024^2,
    `Dense seconds` = run(dense_x, "dense"),
    `Sparse seconds` = run(sparse_x, "sparse"),
    check.names = FALSE
  )
}

benchmark_results <- rbind(
  benchmark_linf(1000),
  benchmark_linf(5000)
)
```

# Introduction

Compositional observations record relative amounts. Classical compositional
data analysis represents strictly positive compositions through log-ratios
[@aitchison1986book]. In R, \CRANpkg{compositions} implements the Aitchison
framework, \CRANpkg{robCompositions} adds robust procedures and several
zero-handling methods, and \CRANpkg{zCompositions} provides methods for
replacing zeros and left-censored values [@compositionsPackage;
@robCompositionsPackage; @zCompositionsPackage]. These packages address a
broad range of estimation and inference problems.

The \CRANpkg{linf} package addresses a different task: describing a sample by
its dominant feature and, when useful, by a hierarchy of successively
subdominant features [@linfPackage]. The construction is defined for
nonnegative matrices and leaves zero coordinates at zero. It provides a
categorical summary whose labels refer directly to observed features. It does
not supply regression, differential-abundance testing, or a probabilistic
model, and it should be used alongside rather than in place of methods suited
to those questions.

Dominance summaries are particularly familiar in vaginal microbiome research.
The original vaginal Community State Types (CSTs) were obtained by clustering
16S rRNA profiles [@ravel2011vaginal]. VALENCIA later provided reproducible
nearest-centroid assignment to fitted CST reference profiles
[@france2020valencia]. Dominant Community State Types (dCSTs) are related but
distinct: their depth-1 labels arise from the largest component, and the
retained state vocabulary is selected by a minimum support threshold. The
geometric construction is developed separately by @gajerRavel2025linf. This
article focuses on its implementation in R.

# Representation and software design

For a nonzero nonnegative row vector $x=(x_1,\ldots,x_p)$,

$$
z = \frac{x}{\max_j x_j}
$$

is its L-infinity-normalized representation. Its largest coordinate is one,
and every zero in $x$ remains zero. The identity of the largest component is
unchanged by the normalization. The package nevertheless keeps normalization
as an explicit step because the normalized matrix is also used by refinement,
landmark, and coordinate functions.

The interface follows four design choices.

1. Samples are rows and features are columns throughout the package.
2. Stable feature identifiers and human-readable feature labels are stored
   separately. This avoids using abbreviated taxonomic names as join keys.
3. Fitted objects retain pure- and absorb-policy assignments. `dcst.view()`
   switches the active view without refitting the hierarchy.
4. Dense matrices and sparse `Matrix` objects use separate backends. The
   default `backend = "auto"` preserves sparse input where the operation can
   remain sparse [@MatrixPackage].

Table \@ref(tab:interface) summarizes the public workflow. The low-level
functions remain available for inspection, while
`linf.dcst.landmark.pipeline()` supplies a compact end-to-end route.

```{r interface, echo=FALSE}
interface <- data.frame(
  Stage = c(
    "Preparation", "Normalization", "Dominance", "Depth 1",
    "Refinement", "Transfer", "Representatives", "Coordinates"
  ),
  Function = c(
    "filter.asv()", "normalize.linf()", "linf.dominant.features()",
    "linf.csts()", "refine.linf.csts()", "transfer.dcsts()",
    "linf.landmarks()", "linf.hypercube.embedding()"
  ),
  Output = c(
    "filtered counts and relatives", "L-infinity relatives",
    "dominant feature IDs and labels", "fitted depth-1 dCST object",
    "updated hierarchy", "assignments to a frozen hierarchy",
    "observed and summary profiles", "bounded reference-component chart"
  )
)
knitr::kable(
  interface,
  caption = "Main stages of the linf interface.",
  booktabs = TRUE
)
```

## Dominance, support, and policy views

`linf.dominant.features()` assigns each nonzero row to an index attaining the
row maximum. The default resolves exact ties by the first maximum in column
order; `tie.method = "random"` and `tie.method = "error"` make alternative
choices explicit. All-zero rows receive `NA` at this stage.

`linf.csts()` counts the resulting depth-1 dominance sample sets. Features
with support of at least `n0` form the retained vocabulary. Under the pure
policy, samples from smaller provisional sets are labeled `RARE_DOMINANT`.
Under the absorb policy, each such sample is reassigned to its largest retained
feature. The latter often gives a more convenient biological summary, but it
is cohort-fitted: changing the cohort can change which features pass `n0`.

```{r toy-workflow}
toy <- rbind(
  s1 = c(A = 8, B = 2, C = 0),
  s2 = c(A = 7, B = 3, C = 0),
  s3 = c(A = 2, B = 9, C = 1),
  s4 = c(A = 1, B = 2, C = 8),
  s5 = c(A = 4, B = 4, C = 0)
)

toy_linf <- normalize.linf(toy)
toy_dominant <- linf.dominant.features(toy_linf)
toy_fit <- linf.csts(
  toy_linf,
  n0 = 2,
  low.freq.policy = "absorb"
)

data.frame(
  sample = rownames(toy),
  dominant = toy_dominant$label,
  dCST = toy_fit$lineage.label
)
```

`refine.linf.csts()` removes the feature used at the current lineage depth and
repeats the support calculation within selected parent lineages. For example,
a lineage led by *Lactobacillus iners* and then *Gardnerella vaginalis* records
an ordered dominance sequence rather than an unordered pair. Automated refinement uses
`refinement.factor` to restrict refinement to sufficiently large parents;
`refine.linf.csts.iter()` also permits an explicit set of lineages.

## Frozen transfer and zero-aware coordinates

Raw dominant-feature assignment is sample-wise once the feature universe,
column order, and tie rule are fixed. A thresholded dCST hierarchy is not:
`n0` and the realized child lineages are fitted from a reference cohort.
`transfer.dcsts()` freezes that hierarchy, aligns a new matrix by feature ID or
label, and walks each new sample through the realized tree. Extra query
features are ignored, and missing reference features are treated as zero.

The package also implements a bounded homogeneous-coordinate chart. Ordinary
ratios to a reference component are undefined when that component is zero.
`linf.hypercube.embedding()` uses finite ratios when possible and a boundary
extension otherwise, returning $p-1$ coordinates in the unit cube.

```{r embedding-example}
coordinate_example <- rbind(
  finite_reference = c(A = 2, B = 1, C = 1),
  zero_reference = c(A = 0, B = 2, C = 1)
)

round(linf.hypercube.embedding(
  coordinate_example,
  reference = "A",
  lambda = log(2)
), 3)
```

# Vaginal microbiome workflow

The bundled `valencia2k` object is a seed-controlled, sub-CST-stratified sample
of 2,000 profiles from the public VALENCIA training data. It contains 178
taxonomic features, the original VALENCIA labels, and read depths. The object
is small enough to rebuild the examples during article rendering. Its purpose
is software demonstration, not estimation of CST prevalence.

```{r vaginal-fit}
data(valencia2k, package = "linf")

z <- normalize.linf(valencia2k$rel)
dominant <- linf.dominant.features(z)
fit <- linf.csts(z, n0 = 30, low.freq.policy = "absorb")
fit2 <- refine.linf.csts(
  z,
  fit,
  n0 = 15,
  refinement.factor = 2,
  low.freq.policy = "absorb",
  verbose = FALSE
)

c(
  observed_dominant_features = length(dominant$observed.levels),
  retained_depth1_states = length(fit$retained.feature.labels),
  realized_depth2_states = length(unique(fit2$lineage.label))
)
```

The 35 observed dominant features reduce to nine retained depth-1 states at
`n0 = 30`; depth-2 refinement yields 28 realized absorb-policy lineages. Figure
\@ref(fig:valencia-concordance) compares the depth-1 assignments with the seven
VALENCIA CST labels. The weighted within-dCST majority fraction is
`r sprintf("%.1f", 100 * valencia_majority_fraction)`%. This number describes
the cross-tabulation and is not an accuracy estimate against a biological gold
standard. VALENCIA uses similarities to fitted reference centroids, whereas a
dCST uses ordered component dominance and a support threshold.

```{r valencia-concordance, echo=FALSE, fig.pos="H", fig.cap="Row percentages for absorb-policy depth-1 dCSTs versus VALENCIA CST assignments in the bundled 2,000-sample vaginal dataset. The constructions correspond closely but are not identical.", fig.alt=output_alt("Heat map with nine dCST rows and seven VALENCIA CST columns. Most rows have one dark cell, indicating that each dCST is concentrated in one VALENCIA CST.")}
pct <- valencia_concordance_pct
row_labels <- vapply(rownames(pct), pretty_taxon, character(1L))
col_labels <- colnames(pct)

op <- par(mar = c(4.2, 11.5, 1.2, 1.5))
image(
  x = seq_len(ncol(pct)),
  y = seq_len(nrow(pct)),
  z = t(pct[nrow(pct):1, , drop = FALSE]),
  col = hcl.colors(20, "Blues", rev = TRUE),
  zlim = c(0, 100),
  axes = FALSE,
  xlab = "VALENCIA CST",
  ylab = ""
)
axis(1, at = seq_len(ncol(pct)), labels = col_labels)
axis(2, at = seq_len(nrow(pct)), labels = rev(row_labels), las = 2, cex.axis = 0.8)
for (i in seq_len(nrow(pct))) {
  for (j in seq_len(ncol(pct))) {
    value <- pct[nrow(pct) - i + 1L, j]
    if (value >= 1) {
      text(j, i, sprintf("%.0f", value), cex = 0.65,
           col = if (value >= 55) "white" else "black")
    }
  }
}
box()
par(op)
```

To demonstrate transfer without refitting, we selected 1,500 profiles with a
fixed seed, fitted a depth-2 absorb-policy hierarchy, and assigned the 500 held
out profiles with `transfer.dcsts()`. The held-out weighted within-dCST majority
fraction relative to VALENCIA CSTs was
`r sprintf("%.1f", 100 * transfer_majority_fraction)`%. The calculation tests
the software's frozen-vocabulary path; it is not an independent clinical
validation because both subsets originate from the same public training
resource.

```{r transfer-example}
transferred <- transfer.dcsts(
  valencia_linf[test_index, ],
  transfer_fit,
  depth = 1:2,
  view = "absorb"
)

head(transferred$assignment, 4)
```

# Gut microbiome and sparse matrices

The bundled `agp_gut` object contains 766 American Gut Project profiles
[@mcdonald2018american]. It is deliberately stratified to retain several
uncommon dCSTs and is not a probability sample. The phenotype columns support
data-alignment demonstrations but must not be used to estimate prevalence or
population associations.

After the documented library-size and prevalence filters, the example retains
`r nrow(gut_filtered$counts)` samples and `r ncol(gut_filtered$counts)`
features. Seven features pass `n0 = 30` under the absorb policy. The median
largest relative abundance is `r sprintf("%.2f", median(gut_dominance))`, much
lower than in strongly single-species-dominated vaginal communities. Figure
\@ref(fig:gut-summary) shows both the retained dCST sizes and the continuous
dominance-strength distribution. The latter should be inspected alongside the
categorical assignments because a dCST label alone does not record how strongly
the named feature dominates.

```{r gut-summary, echo=FALSE, fig.pos="H", fig.cap="Absorb-policy depth-1 dCST sizes (left) and dominance strength (right) in the filtered, stratified American Gut demonstration subset. The subset is intended for software demonstration rather than population inference.", fig.alt=output_alt("Two panels. The left is a bar chart of seven gut dCST sizes, led by Bacteroides. The right is a histogram of the largest relative abundance per sample, concentrated below one half with a right tail.")}
op <- par(mfrow = c(1, 2), mar = c(7.5, 4, 2, 0.5))
bars <- barplot(
  gut_sizes,
  col = "#3B82A0",
  border = NA,
  las = 2,
  cex.names = 0.72,
  ylab = "Samples",
  main = "Depth-1 dCST size"
)
text(bars, gut_sizes, labels = gut_sizes, pos = 3, cex = 0.7)
par(mar = c(4, 4, 2, 0.5))
hist(
  gut_dominance,
  breaks = seq(0, 1, by = 0.04),
  col = "#75B798",
  border = "white",
  xlab = "Largest relative abundance",
  ylab = "Samples",
  main = "Dominance strength"
)
abline(v = median(gut_dominance), lty = 2, lwd = 1.5, col = "#7A1F5C")
par(op)
```

Sparse input matters when a feature universe is large and each sample contains
few observed features. Table \@ref(tab:benchmark-table) reports a small synthetic
benchmark with 1,000 features and 1% nonzero entries. Each time is the median
of five calls that normalize the matrix and fit absorb-policy depth-1 dCSTs.
The values are build-machine measurements rather than general performance
guarantees. In this example the sparse representation reduces object size by
about two orders of magnitude and also lowers elapsed time.

```{r benchmark-table, echo=FALSE}
benchmark_display <- benchmark_results
benchmark_display[, c("Dense MB", "Sparse MB")] <- round(
  benchmark_display[, c("Dense MB", "Sparse MB")], 2
)
benchmark_display[, c("Dense seconds", "Sparse seconds")] <- round(
  benchmark_display[, c("Dense seconds", "Sparse seconds")], 3
)
knitr::kable(
  benchmark_display,
  format = "pipe",
  caption = "Dense and sparse backend benchmark on synthetic matrices with 1% nonzero entries.",
  booktabs = TRUE
)
```

The core package imports only \CRANpkg{Matrix} and `methods`. Plotting,
interactive displays, testing, and vignette generation remain suggested
dependencies. Sparse normalization and dominance assignment do not densify the
input. `linf.hypercube.embedding()` returns a dense matrix by design because
its bounded homogeneous coordinates are generally dense.

# Limitations and appropriate use

Several aspects of a dCST analysis require an explicit scientific decision.

- **Zeros:** the software preserves observed zeros but cannot determine whether
  a zero means biological absence, censoring, or inadequate sampling.
- **Feature universe:** dominance is defined relative to the columns supplied.
  Taxonomic aggregation, filtering, or a different reference database can
  change assignments.
- **Ties and ordering:** exact ties require a declared rule. The default first
  maximum is reproducible only when column order is retained.
- **Support threshold:** `n0` controls the fitted state vocabulary. Sensitivity
  to plausible thresholds should be reported.
- **Transfer:** portable assignment requires a frozen reference hierarchy and
  stable feature identifiers. Recomputing `n0` on a target cohort is refitting,
  not transfer.
- **Interpretation:** dCSTs summarize ordered dominance. They do not prove that
  the underlying composition distribution is clustered, and they do not
  replace continuous analyses when effect estimation is the objective.

The pure policy is useful when the scientific question requires retained states
to contain only samples originally dominated by the named feature. The absorb
policy is easier to display and guarantees assignment when at least one state
is retained, but its reassigned samples may be only weakly dominated by that
feature. The package stores both views so this choice can be inspected rather
than hidden.

# Reproducibility and availability

\CRANpkg{linf} is distributed under the MIT license from CRAN and developed at
<https://github.com/pgajer/linf>. The package includes the two demonstration
datasets used here, source scripts under `data-raw/`, unit tests, and two
installed workflow vignettes. This article is generated from `linf.Rmd` with
the R Journal's \CRANpkg{rjtools} format. The build installs the current package
into an isolated article library, evaluates every result, runs the citation and
journal-format checks, and produces a submission archive. The submission gate
also verifies that the described version is publicly available on CRAN.

# Discussion

The package turns component dominance into a compact R workflow while keeping
the decisions that affect interpretation visible. Raw assignment is
sample-wise; support filtering is cohort-fitted; and transfer freezes the
resulting hierarchy. This distinction is important in applications where a
state vocabulary is learned in one cohort and reused elsewhere.

The vaginal example shows why the representation can be useful: dominant
features recover much of the broad structure encoded by VALENCIA CSTs while
retaining labels tied directly to taxa. The differences are equally important.
VALENCIA considers similarity to complete reference profiles, whereas dCSTs
consider ordered dominance within a retained vocabulary. In more even
communities, such as the gut demonstration, dominance strength provides
essential context and a categorical label may be less informative.

Future package development should emphasize diagnostics for threshold
sensitivity, summaries of transfer uncertainty, and methods for comparing
frozen hierarchies across cohorts. Such diagnostics can be added without
changing the central separation between feature identity, fitted state
vocabulary, and sample assignment.

# Acknowledgments

Generative AI tools were used for editorial assistance and software-supported
quality checks during preparation of this article. The author reviewed and
takes responsibility for the scientific interpretation, citations, code, and
final text.
