---
title: "Exploratory Gut DCST-Phenotype Analysis"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 9,
  fig.height = 6,
  eval = FALSE
)
```

This companion article presents the full 5,000-sample American Gut Project
(AGP) DCST analysis that underlies the curated subset in
`vignette("linf-intro")`. Unlike the package vignette, this article is *not*
built during `R CMD check`---it uses external data files and takes several
minutes to run. The code below is fully reproducible given access to the
PRIME-processed AGP abundance table.

The association calculations are exploratory and unadjusted. They do not
establish causal, diagnostic, or clinically validated relationships.

## Data

The analysis uses the PRIME-processed SILVA species-level abundance table
for PRJEB11419 (American Gut Project, ~26,900 gut samples) and the
accompanying sample metadata. These files are produced by the
[PRIME pipeline](https://github.com/pgajer/PRIME). Their locations are supplied
through environment variables rather than maintainer-specific paths.

```{r data-paths}
library(linf)

abundance_file <- Sys.getenv("LINF_AGP_ABUNDANCE")
metadata_file  <- Sys.getenv("LINF_AGP_METADATA")
if (!nzchar(abundance_file) || !nzchar(metadata_file)) {
  stop("Set LINF_AGP_ABUNDANCE and LINF_AGP_METADATA before rendering.")
}
```

## Phase 1: Data Loading and Subsampling

```{r load-data}
## Load full abundance table
counts_full <- read.csv(gzfile(abundance_file), row.names = 1,
                         check.names = FALSE)
counts_full <- as.matrix(counts_full)
storage.mode(counts_full) <- "integer"
cat("Full abundance table:", nrow(counts_full), "samples x",
    ncol(counts_full), "taxa\n")

## Load metadata and filter to AGP
meta_full <- read.csv(gzfile(metadata_file), stringsAsFactors = FALSE)
agp_meta  <- meta_full[meta_full$BioProject == "PRJEB11419", ]
cat("AGP metadata:", nrow(agp_meta), "samples\n")

## Align to abundance table
common <- intersect(rownames(counts_full), agp_meta$Run)
counts_agp <- counts_full[common, ]
agp_meta   <- agp_meta[match(common, agp_meta$Run), ]
cat("Matched:", nrow(counts_agp), "samples\n")
```

```{r subsample}
## Subsample 5,000 samples with adequate library size
set.seed(42L)
lib_sizes <- rowSums(counts_agp)
good_idx  <- which(lib_sizes >= 1000)
cat("Samples with lib >= 1000:", length(good_idx), "\n")

sub_idx    <- sort(sample(good_idx, min(5000, length(good_idx))))
counts_sub <- counts_agp[sub_idx, ]
meta_sub   <- agp_meta[sub_idx, ]
cat("Subsampled:", nrow(counts_sub), "samples\n")
```

## Phase 2: QC and DCST Construction

```{r filter}
filt <- filter.asv(counts_sub, min.lib = 1000, prev.prop = 0.05,
                   min.count = 2)
cat("After filtering:", nrow(filt$counts), "samples x",
    ncol(filt$counts), "taxa\n")
```

```{r dcst-pipeline}
## L-infinity normalization
M <- normalize.linf(filt$counts)

## Depth-1 DCSTs
csts1 <- linf.csts(M, n0 = 50)
cat("\nDepth-1 DCSTs:\n")
print(sort(table(csts1$cell.label), decreasing = TRUE))

## Depth-2 refinement
csts2 <- refine.linf.csts(M, csts1, n0 = 50)
cat("\nDepth-2 DCSTs (n >= 50):\n")
tab2 <- sort(table(csts2$cell.label), decreasing = TRUE)
print(tab2[tab2 >= 50])
```

```{r dcst-barplot, fig.width = 12, fig.height = 6}
tab1 <- sort(table(csts1$cell.label), decreasing = TRUE)
par(mar = c(12, 4, 3, 1))
barplot(tab1,
        col = ifelse(names(tab1) == "RARE_DOMINANT", "#e74c3c", "#3498db"),
        las = 2, cex.names = 0.8,
        ylab = "Number of samples",
        main = "Gut DCST Size Distribution (AGP, n=5000, depth 1)")
```

## Phase 3: Exploratory Phenotype Association Testing

```{r parse-metadata}
## Parse disease indicators from AGP Phenotype field
conditions <- c(
  IBS = "irritable bowel syndrome",
  IBD = "inflammatory bowel disease",
  Diabetes = "diabetes",
  Autoimmune = "autoimmune",
  Seasonal_allergies = "seasonal allergies",
  Migraine = "migraine",
  Acid_reflux = "acid reflux",
  Lung_disease = "lung disease",
  Cardiovascular_disease = "cardiovascular disease",
  Skin_condition = "skin condition"
)

## Build disease indicator matrix aligned to filtered samples
kept_runs <- rownames(M)
meta_aligned <- meta_sub[match(kept_runs, meta_sub$Run), ]

disease_mat <- matrix(0L, nrow = nrow(meta_aligned), ncol = length(conditions),
                      dimnames = list(kept_runs, names(conditions)))
for (i in seq_along(conditions)) {
  pheno <- tolower(as.character(meta_aligned$Phenotype))
  disease_mat[, i] <- as.integer(grepl(conditions[i], pheno, fixed = TRUE))
}

## Add obesity from BMI
bmi <- suppressWarnings(as.numeric(meta_aligned$Host_BMI))
disease_mat <- cbind(disease_mat, Obesity = as.integer(!is.na(bmi) & bmi >= 30))
```

```{r fisher-tests}
## Fisher's exact test for each DCST x condition pair
dcst_labels <- csts1$cell.label
kept_dcsts  <- names(tab1[tab1 >= 30])

results <- data.frame()

for (dcst in kept_dcsts) {
  in_dcst <- dcst_labels == dcst

  for (j in seq_len(ncol(disease_mat))) {
    cond <- colnames(disease_mat)[j]
    d <- disease_mat[, j]

    a <- sum(in_dcst & d == 1)
    b <- sum(in_dcst & d == 0)
    cc <- sum(!in_dcst & d == 1)
    dd <- sum(!in_dcst & d == 0)

    ft <- fisher.test(matrix(c(a, b, cc, dd), nrow = 2))

    ## Woolf CI for log-OR
    if (all(c(a, b, cc, dd) > 0)) {
      se <- sqrt(1/a + 1/b + 1/cc + 1/dd)
      ci <- exp(log(ft$estimate) + c(-1, 1) * 1.96 * se)
    } else {
      ci <- c(NA, NA)
    }

    results <- rbind(results, data.frame(
      DCST = dcst, Condition = cond,
      n_DCST = sum(in_dcst), n_cases = a,
      prev_DCST = round(a / sum(in_dcst), 3),
      prev_pop  = round((a + cc) / length(dcst_labels), 3),
      OR = round(ft$estimate, 2),
      CI_low = round(ci[1], 2), CI_high = round(ci[2], 2),
      p_value = ft$p.value,
      stringsAsFactors = FALSE
    ))
  }
}

## BH correction
results$q_value <- p.adjust(results$p_value, method = "BH")
results <- results[order(results$p_value), ]
rownames(results) <- NULL

cat("Total tests:", nrow(results), "\n")
cat("Exploratory q < 0.05 signals:", sum(results$q_value < 0.05), "\n")
```

```{r top-results}
## Display top 20 associations
top20 <- head(results, 20)
top20$p_value <- formatC(top20$p_value, format = "e", digits = 2)
top20$q_value <- formatC(top20$q_value, format = "e", digits = 2)
knitr::kable(top20, caption = "Top 20 exploratory DCST-phenotype results by p-value",
             row.names = FALSE)
```

```{r heatmap, fig.width = 10, fig.height = 8}
## Heatmap of exploratory signals
sig <- results[results$q_value < 0.2, ]
if (nrow(sig) > 0) {
  or_mat <- tapply(sig$OR, list(sig$DCST, sig$Condition), identity)
  or_mat[is.na(or_mat)] <- 1  # neutral OR

  log2_or <- log2(or_mat)
  log2_or[log2_or > 3] <- 3
  log2_or[log2_or < -3] <- -3

  ## Simple heatmap
  image(t(log2_or), col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(50),
        axes = FALSE, main = "Exploratory DCST x Phenotype Results (q < 0.2)")
  axis(1, at = seq(0, 1, length.out = nrow(log2_or)),
       labels = rownames(log2_or), las = 2, cex.axis = 0.7)
  axis(2, at = seq(0, 1, length.out = ncol(log2_or)),
       labels = colnames(log2_or), las = 1, cex.axis = 0.8)
}
```

## Phase 4: Landmark Analysis

```{r landmarks}
## Landmark points for depth-1 DCSTs
lm1 <- linf.landmarks(M, csts1, depth = 1,
                       landmark.types = c("endpoint.max", "endpoint.min",
                                          "mean.rep"))

## Focus on disease-associated DCSTs
disease_dcsts <- unique(sig$DCST[sig$q_value < 0.1])

for (dcst_name in disease_dcsts) {
  if (!dcst_name %in% names(lm1)) next
  cat("\n===", dcst_name, "===\n")
  lm_dcst <- lm1[[dcst_name]]

  for (ltype in c("endpoint.max", "mean.rep")) {
    if (!ltype %in% names(lm_dcst)) next
    cat(ltype, ": sample", lm_dcst[[ltype]]$sample, "\n")
    profile <- lm_dcst[[ltype]]$profile
    top5 <- head(sort(profile, decreasing = TRUE), 5)
    for (i in seq_along(top5)) {
      cat(sprintf("  %-40s %.3f\n", names(top5)[i], top5[i]))
    }
  }
}
```

## Descriptive and Exploratory Summary

Using 5,000 AGP gut samples, the L-infinity DCST pipeline identified:

- **14 depth-1 DCSTs** (n0 = 50), dominated by *Bacteroides* (~35%) and
  *Escherichia-Shigella* (~23%).
- **12.8% of samples in RARE_DOMINANT**, representing the tail of unusual
  single-species dominance.
- **11 exploratory DCST-disease signals** with q < 0.05 after
  Benjamini-Hochberg correction in this analysis cohort.

The strongest signals include *Prevotella_7* enrichment in obesity (OR ~ 4.0),
*Pasteurellaceae* enrichment in cardiovascular disease (OR ~ 8.8),
*Prevotella_9* depletion in autoimmune conditions (OR ~ 0.3), and
*Escherichia-Shigella* enrichment in IBS (OR ~ 1.5).

These unadjusted results show how DCST labels can be carried into a downstream
association workflow. They require covariate adjustment, sensitivity analyses,
and independent validation before biological or clinical interpretation.

## Session Info

```{r session-info}
sessionInfo()
```
