---
title: "VALENCIA Four-Component Hypercube Embeddings"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 6)
library(linf)
if (!requireNamespace("plotly", quietly = TRUE) ||
    !requireNamespace("htmltools", quietly = TRUE)) {
  stop("This website article requires plotly and htmltools.")
}
data(valencia_linf_hypercube_1k, package = "linf")
X4 <- valencia_linf_hypercube_1k$rel4
meta <- valencia_linf_hypercube_1k$meta
stopifnot(nrow(X4) == nrow(meta), !anyDuplicated(meta$sample_id))
component.colors <- c(Li = "#0072B2", Lc = "#E69F00", Gv = "#009E73", Bv = "#CC79A7")
embeddings <- lapply(colnames(X4), function(ref) linf.hypercube.embedding(X4, reference = ref))
names(embeddings) <- colnames(X4)
```

This **website-only article** shows zero-aware hypercube coordinates for
`r nrow(X4)` samples from the bundled VALENCIA four-component dataset.
The first views use the three coordinates returned by
`linf.hypercube.embedding()` directly. Rotate them to inspect where samples
lie relative to the selected reference component; hover to read the sample,
dominant component, original VALENCIA label and retained abundance mass.

These are coordinate representations of samples from the VALENCIA training
source, not an independent evaluation of classification or biological groups.
Reference choice, selected components and radial scale affect the geometry.
For static installed workflows and provenance, see the
[dataset guide](example-datasets.html); for fitting and transfer, see the
[function guide](function-guide.html).

## Data and coordinate meanings

```{r component-map, echo=FALSE}
map <- valencia_linf_hypercube_1k$component_map
knitr::kable(data.frame(Component = names(map), Feature = unname(map)))
```

The input keeps these four components and divides each sample by their sum.
`selected_mass` records their original mass before that renormalization.
A point's color names the largest of the four selected components, which need
not be the largest feature in the original full profile. VALENCIA CST labels
are retained as hover information; they are not mapped to marker symbols.
All points use a circle; the category table provides counts without requiring hover.

```{r data-summary, echo=FALSE}
knitr::kable(as.data.frame(table(meta$dominant_component)),
             col.names = c("Dominant selected component", "Samples"))
knitr::kable(as.data.frame(table(meta$Val_CST)),
             col.names = c("Original VALENCIA CST", "Samples"))
```

For a positive reference, the chart preserves the direction of the other
three components and maps their combined reference ratio to a radius in
[0, 1]. With reference zero it uses the L-infinity boundary extension.
Each axis below names its non-reference component's **mapped coordinate**,
not its relative abundance. Each reference has its own automatically fitted
radial scale; distances across different charts are not on a common scale.

## A static preview

This projection shows the first two mapped coordinates for reference Li;
the Bv coordinate is omitted here. The interactive chart below adds that
third coordinate. The preview remains usable without interactive graphics.

```{r static-preview, echo=FALSE, fig.cap="Two-coordinate projection of the Li-reference chart for all 1,000 bundled samples. Color identifies the dominant selected component; the third mapped coordinate is omitted."}
preview <- embeddings[["Li"]]
plot(preview[, 1], preview[, 2], pch = 16, cex = 0.6,
     col = grDevices::adjustcolor(component.colors[as.character(meta$dominant_component)], alpha.f = 0.5),
     xlab = "Lc mapped coordinate", ylab = "Gv mapped coordinate",
     xlim = c(0, 1), ylim = c(0, 1), main = "Reference Li: a two-coordinate projection")
legend("topleft", legend = names(component.colors), col = component.colors, pch = 16,
       bty = "n", title = "Dominant selected component")
```

```{r plotting-helper, include=FALSE}
plot_embedding <- function(coords, title, subtitle, axis.labels = colnames(coords), unit.cube = FALSE) {
  stopifnot(ncol(coords) == 3L, nrow(coords) == nrow(meta), all(is.finite(coords)))
  df <- data.frame(x = coords[, 1], y = coords[, 2], z = coords[, 3],
                   dominant_component = factor(meta$dominant_component, levels = names(component.colors)))
  hover <- paste0("sample: ", meta$sample_id,
                  "<br>dominant selected component: ", meta$dominant_component,
                  "<br>VALENCIA CST: ", meta$Val_CST,
                  "<br>selected mass: ", signif(meta$selected_mass, 4))
  p <- plotly::plot_ly(df, x = ~x, y = ~y, z = ~z,
                       color = ~dominant_component, colors = component.colors,
                       text = hover, hoverinfo = "text", type = "scatter3d",
                       mode = "markers", height = 550, marker = list(size = 3, opacity = 0.7, symbol = "circle"))
  axis <- function(label) {
    if (unit.cube) list(title = label, range = c(0, 1)) else list(title = label)
  }
  plotly::layout(p, title = list(text = paste0(title, "<br><sup>", subtitle, "</sup>")),
                  scene = list(xaxis = axis(axis.labels[[1]]), yaxis = axis(axis.labels[[2]]),
                               zaxis = axis(axis.labels[[3]]), aspectmode = "cube",
                               camera = list(eye = list(x = 1.7, y = 1.7, z = 1.7))),
                  legend = list(orientation = "h"), margin = list(t = 90))
}
```

## Native three-coordinate charts

```{r chart-example, eval=FALSE}
library(linf)
X4 <- valencia_linf_hypercube_1k$rel4
emb <- linf.hypercube.embedding(X4, reference = "Li")
```

Axis labels use the component abbreviations from the table; every axis is a
mapped coordinate on [0, 1].

All four component groups and all seven VALENCIA CST categories are retained.
The legend filters dominant-component groups; hover supplies CST details.
Drag to rotate, scroll to zoom, and use the plot toolbar to reset the view.

```{r hypercube-views, echo=FALSE}
native.widgets <- lapply(names(embeddings), function(ref) {
  emb <- embeddings[[ref]]
  scale.text <- paste0("log(lambda) = ", signif(attr(emb, "log.lambda"), 4),
                       "; ", nrow(emb), " samples; native mapped coordinates")
  plot_embedding(emb, paste("Reference", ref), scale.text,
                  axis.labels = attr(emb, "other.labels"),
                  unit.cube = TRUE)
})
htmltools::tagList(native.widgets)
```

## Optional distance-based layouts

The native charts above require no distance embedding or graph optimizer.
Classical multidimensional scaling (MDS) of the original four-component input
is shown separately below. Its axes are rotated distance coordinates, not
component abundances. Graph-based layouts are an additional experiment and
can change distances; they do not validate separation seen in these charts.

<details>
<summary>Show the original four-component data in three MDS coordinates</summary>

```{r original-mds}
mds.coords <- stats::cmdscale(stats::dist(X4), k = 3)
colnames(mds.coords) <- paste("MDS coordinate", 1:3)
plot_embedding(mds.coords, "Original four-component input: classical MDS",
                 "Euclidean distances between four-component compositions")
```

</details>

### The graph used for an optional layout

We use the **union** of directed nearest-neighbor relations: keep an undirected
edge if either endpoint selects the other. All neighbors tied at the kth
distance are included, so ties may give more than k neighbors and row order
does not discard valid edges. Endpoint pairs are canonicalized and deduplicated
after collecting all relations. Zero-distance edges receive machine epsilon
as their positive optimizer length; other edge lengths are Euclidean distances
in the source coordinates.

```{r graph-helpers, include=FALSE}
knn_edges <- function(X, k = 8L) {
  X <- as.matrix(X)
  if (!is.numeric(X) || !nrow(X) || any(!is.finite(X)) ||
      !is.numeric(k) || length(k) != 1L || !is.finite(k) || k < 1 || k %% 1 != 0) {
    stop("knn_edges: use finite numeric points and an integer k >= 1")
  }
  n <- nrow(X)
  if (n == 1L) return(list(edges = matrix(integer(), 0L, 2L), weights = numeric()))
  D <- as.matrix(stats::dist(X))
  if (any(!is.finite(D))) stop("knn_edges: distances exceed numeric range; rescale the points")
  diag(D) <- Inf
  k <- min(k, n - 1L)
  directed <- lapply(seq_len(n), function(i) {
    cutoff <- sort(D[i, ], partial = k)[[k]]
    neighbors <- which(D[i, ] <= cutoff)
    cbind(pmin(i, neighbors), pmax(i, neighbors))
  })
  edges <- unique(do.call(rbind, directed))
  edges <- edges[order(edges[, 1], edges[, 2]), , drop = FALSE]
  list(edges = unname(edges), weights = pmax(D[edges], .Machine$double.eps))
}

graph_components <- function(n, edges) {
  adjacent <- vector("list", n)
  for (i in seq_len(nrow(edges))) {
    a <- edges[i, 1]; b <- edges[i, 2]
    adjacent[[a]] <- c(adjacent[[a]], b)
    adjacent[[b]] <- c(adjacent[[b]], a)
  }
  component <- integer(n)
  count <- 0L
  for (root in seq_len(n)) if (component[[root]] == 0L) {
    count <- count + 1L
    pending <- root
    component[[root]] <- count
    while (length(pending)) {
      current <- pending[[length(pending)]]
      pending <- pending[-length(pending)]
      unseen <- adjacent[[current]][component[adjacent[[current]]] == 0L]
      component[unseen] <- count
      pending <- c(pending, unseen)
    }
  }
  component
}
```

```{r graph-summary, echo=FALSE}
graphs <- lapply(embeddings, knn_edges, k = 8L)
graph.summary <- do.call(rbind, lapply(names(graphs), function(ref) {
  graph <- graphs[[ref]]
  component <- graph_components(nrow(X4), graph$edges)
  data.frame(Reference = ref, Samples = nrow(X4), Edges = nrow(graph$edges),
              Components = length(unique(component)), Largest = max(table(component)))
}))
knitr::kable(graph.summary, row.names = FALSE)
```

Components are disconnected sets in the graph. A local graph objective cannot
determine their relative positions; any disconnected layout retains dependence
on its initialization. The table reports this rather than assuming connectivity.
The corrected union graph differs from older article versions that discarded
some one-way relations; an optional layout from it is a new computation.

<details>
<summary>Reproduce the optional Li-reference graph layout with grip</summary>

This optional experiment is off in ordinary website builds. Install a compatible
`grip` version that exports `edge.kk`, set
`LINF_RUN_GRIP=true`, and render this article to run it. It uses the native
Li-reference coordinates as initialization, k = 8 with all cutoff ties,
one stage of at most 20 optimizer iterations and seed 1. Keep its diagnostics when comparing
layouts; a rendered optimizer result is not a convergence guarantee.

```{r optional-grip}
if (identical(Sys.getenv("LINF_RUN_GRIP"), "true")) {
  if (!requireNamespace("grip", quietly = TRUE) ||
      !"edge.kk" %in% getNamespaceExports("grip")) {
    stop("This optional experiment needs a compatible grip installation.")
  }
  optimize <- getExportedValue("grip", "edge.kk")
  graph <- graphs[["Li"]]
  layout.result <- optimize(coords = embeddings[["Li"]], edges = graph$edges,
                             n = nrow(X4), edge.weights = graph$weights, dim = 3L,
                             max.iter = 20L, density.mix.schedule = 1, return.trace = FALSE,
                             diagnostics = TRUE, seed = 1L)
  print(layout.result$metadata$stage_summaries)
  layout.coords <- as.matrix(layout.result$coords)
  colnames(layout.coords) <- paste("Layout coordinate", 1:3)
  plot_embedding(layout.coords, "Li chart: optional graph layout",
                   "Native initialization; union 8-nearest-neighbor graph, including ties")
} else {
  cat("Optional grip optimization was not run. Native charts and graph diagnostics are available above.")
}
```

</details>

## Reproducibility notes

The bundled object contains four-component abundances, not precomputed charts.
`data-raw/build_valencia_linf_hypercube_1k.R` selects Li, Lc, Gv and Bv from the
VALENCIA 13k training matrix, drops zero selected-mass rows, L1-normalizes,
stratifies by the dominant selected component, and samples 1,000 rows with
`set.seed(20261604)`. The charts, static preview, category counts and graph
diagnostics here are generated at render time. No source-data download or
optional optimizer is required for the ordinary article build.
