---
title: "VALENCIA Four-Component Hypercube Embeddings"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

This lightweight package article demonstrates the zero-aware homogeneous
coordinate embedding implemented by `linf.hypercube.embedding()` on a bundled
VALENCIA-derived example dataset. The source object,
`valencia_linf_hypercube_1k`, is generated by
`data-raw/build_valencia_linf_hypercube_1k.R` from the VALENCIA 13k training
set. It keeps four phylotype coordinates,

```{r component-map}
library(linf)

candidate_roots <- normalizePath(c(file.path("..", ".."), "."), mustWork = FALSE)
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (!exists("linf.hypercube.embedding", mode = "function") &&
    requireNamespace("pkgload", quietly = TRUE) &&
    !is.na(pkg_root)) {
  pkgload::load_all(pkg_root, quiet = TRUE)
}

local_data_path <- file.path("..", "..", "data", "valencia_linf_hypercube_1k.rda")
if (file.exists(local_data_path)) {
  load(local_data_path)
} else if (file.exists(file.path("data", "valencia_linf_hypercube_1k.rda"))) {
  load(file.path("data", "valencia_linf_hypercube_1k.rda"))
} else {
  data(valencia_linf_hypercube_1k, package = "linf")
}

valencia_linf_hypercube_1k$component_map
```

and L1-normalizes each sample over these four selected components. The resulting
matrix is four-dimensional and compositional. For each reference component, the
hypercube embedding gives a three-dimensional zero-aware homogeneous coordinate
chart.

```{r data-summary}
X4 <- valencia_linf_hypercube_1k$rel4
meta <- valencia_linf_hypercube_1k$meta

dim(X4)
round(colSums(X4) / nrow(X4), 3)
table(meta$dominant_component)
table(meta$Val_CST)
```

## Embedding Helpers

The figures below use two interactive three-dimensional views.

1. Classical MDS embeds the Euclidean distance matrix into three dimensions.
2. If the optional `grip` package is available, edge-KK starts from the MDS
   coordinates and refines the layout using a local k-nearest-neighbor graph
   whose edge lengths are measured in the source coordinate system.

```{r helpers}
if (!requireNamespace("plotly", quietly = TRUE)) {
  stop("This article preview requires the plotly package.")
}
if (!requireNamespace("htmltools", quietly = TRUE)) {
  stop("This article preview requires the htmltools package.")
}

has_grip <- requireNamespace("grip", quietly = TRUE)
edge_kk_fun <- NULL
if (has_grip && exists("grip.optimize.edge.kk.layout", envir = asNamespace("grip"))) {
  edge_kk_fun <- getFromNamespace("grip.optimize.edge.kk.layout", "grip")
}

mds3 <- function(X) {
  d <- stats::dist(X)
  fit <- stats::cmdscale(d, k = 3, eig = TRUE)
  coords <- as.matrix(fit$points)
  colnames(coords) <- c("Dim1", "Dim2", "Dim3")
  coords
}

knn_edges <- function(X, k = 8L) {
  D <- as.matrix(stats::dist(X))
  n <- nrow(D)
  edges <- vector("list", n)
  weights <- vector("list", n)

  for (i in seq_len(n)) {
    ord <- order(D[i, ], decreasing = FALSE)
    nbrs <- ord[ord != i][seq_len(min(k, n - 1L))]
    keep <- nbrs > i
    if (any(keep)) {
      nbrs <- nbrs[keep]
      edges[[i]] <- cbind(i, nbrs)
      weights[[i]] <- D[cbind(i, nbrs)]
    }
  }

  edge_matrix <- do.call(rbind, edges[lengths(edges) > 0L])
  edge_weights <- unlist(weights[lengths(weights) > 0L], use.names = FALSE)

  list(edges = edge_matrix, weights = pmax(edge_weights, .Machine$double.eps))
}

edge_kk3 <- function(X, initial, k = 8L, max_iter = 20L) {
  if (is.null(edge_kk_fun)) {
    return(NULL)
  }

  graph <- knn_edges(X, k = k)
  out <- edge_kk_fun(
    coords = initial,
    edges = graph$edges,
    n = nrow(X),
    edge_weights = graph$weights,
    dim = 3L,
    max_iter = max_iter,
    return_trace = FALSE,
    diagnostics = FALSE,
    seed = 1L
  )

  coords <- as.matrix(out$coords)
  colnames(coords) <- c("Dim1", "Dim2", "Dim3")
  coords
}

plot_embedding <- function(coords, title, subtitle) {
  df <- data.frame(
    coords,
    dominant_component = meta$dominant_component,
    Val_CST = meta$Val_CST,
    selected_mass = meta$selected_mass,
    sample_id = meta$sample_id,
    stringsAsFactors = FALSE
  )

  hover <- paste0(
    "sample: ", df$sample_id,
    "<br>dominant: ", df$dominant_component,
    "<br>Val_CST: ", df$Val_CST,
    "<br>selected mass: ", signif(df$selected_mass, 4)
  )

  p <- plotly::plot_ly(
    df,
    x = ~Dim1,
    y = ~Dim2,
    z = ~Dim3,
    color = ~dominant_component,
    symbol = ~Val_CST,
    text = hover,
    hoverinfo = "text",
    type = "scatter3d",
    mode = "markers",
    marker = list(size = 3, opacity = 0.75)
  )

  plotly::layout(
    p,
    title = list(text = paste0(title, "<br><sup>", subtitle, "</sup>")),
    scene = list(
      xaxis = list(title = "Dim 1"),
      yaxis = list(title = "Dim 2"),
      zaxis = list(title = "Dim 3")
    ),
    legend = list(orientation = "h")
  )
}

embedding_panel <- function(label, X, reference = NULL) {
  title <- if (is.null(reference)) {
    paste0(label, ": 4D composition")
  } else {
    paste0(label, ": reference ", reference)
  }

  mds_coords <- mds3(X)
  widgets <- list(
    htmltools::tags$h3(paste0(title, " -- classical MDS")),
    plot_embedding(
      mds_coords,
      title = paste0(title, " -- MDS"),
      subtitle = "Classical MDS of Euclidean distances in the source coordinates."
    )
  )

  kk_coords <- edge_kk3(X, mds_coords)
  if (!is.null(kk_coords)) {
    widgets <- c(
      widgets,
      list(
        htmltools::tags$h3(paste0(title, " -- edge-KK")),
        plot_embedding(
          kk_coords,
          title = paste0(title, " -- edge-KK"),
          subtitle = "MDS initialization refined on an 8-nearest-neighbor graph with source-space edge lengths."
        )
      )
    )
  } else {
    widgets <- c(
      widgets,
      list(htmltools::tags$p("Optional edge-KK layout skipped because the grip package is not available."))
    )
  }

  htmltools::tagList(widgets)
}
```

## Four-Component Composition

This is the original four-component compositional dataset before choosing a
homogeneous-coordinate reference component.

```{r original-4d-view}
embedding_panel("VALENCIA four-component data", X4)
```

## Hypercube Embeddings

The four panels below show the zero-aware hypercube chart obtained by taking
each component in turn as the reference coordinate.

```{r hypercube-views}
panels <- lapply(colnames(X4), function(ref) {
  emb <- linf.hypercube.embedding(X4, reference = ref)
  embedding_panel("Hypercube embedding", emb, reference = ref)
})

htmltools::tagList(panels)
```

## Reproducibility Notes

The bundled dataset is intentionally small enough to ship with `linf`. It is
created from the VALENCIA 13k source matrix by:

1. selecting `Li`, `Lc`, `Gv`, and `Bv`;
2. dropping rows with zero total mass in these four selected components;
3. L1-normalizing over the selected components;
4. stratifying by the dominant selected component; and
5. sampling 1,000 rows with `set.seed(20261604)`.

The four hypercube charts are generated at render time using
`linf.hypercube.embedding()`, so this article previews both the bundled data and
the exported embedding API.
